---
title: "cluster_v3"
output: html_document
---

# Preanalysis

## Set inputs
```{r, results='hide'}
library(DESeq2)
library(dplyr)
# library(egg)
# library(edgeR)
library(ggplot2)
library(ggrepel)
# library(limma)
# library(matrixStats)
library(tibble)
library(vegan)
library(viridis)

WORKING.DIR="Z:/mchung/leopoldo_segal_07/"
```

## Load R objects
```{r}
load(paste0(WORKING.DIR,"/upstream.R"))
load(paste0(WORKING.DIR,"/de_analysis.R"))
source("Z:/mchung/pipelines/SGSlab_metagenomics/scripts/cluster_functions.R")
```

```{r}
metadata.BAL <- metadata[metadata$sample_type == "BAL",]
metadata.BAL <- metadata.BAL[!(duplicated(metadata.BAL$subject_id)),]

metadata.TA <- metadata[metadata$sample_type == "TA",]
metadata.TA <- metadata.TA[!(duplicated(metadata.TA$subject_id)),]

metadata.UA <- metadata[metadata$sample_type == "UA",]
metadata.UA <- metadata.UA[!(duplicated(metadata.UA$subject_id)),]
```


# Plot dates of all samples

```{r,fig.height=11,fig.width=11}
plot.df <- as.data.frame(matrix(nrow=nrow(metadata),
                                ncol=0))
plot.df$patient <- factor(metadata$subject_id,levels=rev(as.numeric(as.character(unique(metadata$subject_id)))))
plot.df$patient <- factor(metadata$study_id)
plot.df$sample_type <- as.factor(metadata$sample_type)
plot.df$days_post_intub <- as.numeric(as.character(as.Date(metadata$collection_date) - as.Date(metadata$intubation_date)))
plot.df$admit_to_intub <- as.numeric(as.character((as.Date(metadata$date_hospital_adm) - as.Date(metadata$intubation_date))))
plot.df$intub_to_discharge <- as.numeric(as.character((as.Date(metadata$date_discharge) - as.Date(metadata$intubation_date))))
plot.df$dead_or_alive <- as.factor(metadata$dead_or_alive[match(plot.df$patient,metadata$study_id)])

p1a <- ggplot(plot.df,aes(x=days_post_intub,y=patient,color=sample_type,shape=sample_type))+
  geom_segment(aes(x=admit_to_intub,xend=intub_to_discharge,y=patient),color="grey")+
  geom_vline(aes(xintercept=0),color="red",lty="dashed")+
  facet_grid(dead_or_alive~ .,scales="free_y",space="free_y")+
  scale_color_manual(values=c("grey","blue","purple","orange","darkgrey","green","red"))+
  scale_shape_manual(values=c(17,17,17,17,15,16,16))+
  geom_point(aes(x=admit_to_intub,y=patient,color="Admission",shape="Admission"))+
  geom_point(aes(x=intub_to_discharge,y=patient,color=dead_or_alive,shape=dead_or_alive))+
  geom_point()+
  labs(x="days post intubation",y="patient",color="Event",shape="Event")+
  coord_cartesian(xlim=c(-40,100))+
  scale_x_continuous(expand=c(0,0))+
  theme_bw()+
  theme(panel.grid.major.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        legend.position="bottom",
        strip.text = element_blank(),
        strip.background = element_blank())

p1b <- ggplot(plot.df,aes(x=days_post_intub,y=patient,color=sample_type,shape=sample_type))+
  geom_segment(aes(x=admit_to_intub,xend=intub_to_discharge,y=patient),color="grey")+
  geom_vline(aes(xintercept=0),color="red",lty="dashed")+
  facet_grid(dead_or_alive~ .,scales="free_y",space="free_y")+
  scale_x_continuous(breaks=c(100,200,300),expand=c(0,0))+
  scale_color_manual(values=c("grey","blue","purple","orange","darkgrey","green","red"))+
  scale_shape_manual(values=c(17,17,17,17,15,16,16))+
  geom_point(aes(x=admit_to_intub,y=patient,color="Admission",shape="Admission"))+
  geom_point(aes(x=intub_to_discharge,y=patient,color=dead_or_alive,shape=dead_or_alive))+
  geom_point()+
  guides(color='none',shape='none')+
  labs(x="days post intubation",y="patient",color="Event",shape="Event")+
  coord_cartesian(xlim=c(100,325))+
  theme_bw()+
  theme(panel.grid.major.y = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        legend.position="bottom")

pdf(paste0(WORKING.DIR,"/plots/SupFig1.pdf"),
    height=11,
    width=11,
    onefile = F)
egg::ggarrange(p1a,p1b,
               nrow=1,
               widths=c(5,1))
dev.off()
```

# Plot Bray-Curtis PCoA

```{r,fig.height=6,fig.width=6}
counts <- counts$Bracken

sample_types <- list("BAL_and_TA" = c("BAL","TA"),
                     "UA" = c("UA"))
# metadata$weekpostintub <- ifelse(metadata$weekpostintub >=3, "3+",metadata$weekpostintub)

relab <- as.data.frame(apply(counts,2,function(x){x/sum(x)*100}))

plot <- plot_braycurtis_pcoa(counts,
                             metadata$sample_type,
                             title="All Samples\n")

p2 <- plot+
  scale_color_manual(values=c("darkgrey","blue","purple","orange"))+
  theme(plot.title = element_text(size=8))+
  labs(title="All Samples\nPr(>F) = 0.001")

pdf(paste0(WORKING.DIR,"/plots/SupFig2.pdf"),
    height=6,
    width=6)
plot(p2)
dev.off()
```

# Plot sample distribution

## Set single sample list

```{r}
one_sample_per_subject <- function(metadata,sample_types, week){
  metadata <- metadata[metadata$sample_type %in% sample_types,]
  metadata <- metadata[metadata$weekpostintub %in% week,]
  metadata <- metadata[!duplicated(metadata$subject_id),]
  return(metadata$Sample.ID)
}

# samples <- list("BAL_and_TA | Week 1" = one_sample_per_subject(metadata, c("BAL","TA"), week = c(1)),
#                 "BAL_and_TA | Week 2" = one_sample_per_subject(metadata, c("BAL","TA"), week = c(2)),
#                 "BAL_and_TA | Week 3" = one_sample_per_subject(metadata, c("BAL","TA"), week = c(3)),
#                 "BAL_and_TA | Week 4" = one_sample_per_subject(metadata, c("BAL","TA"), week = c(4)),
#                 "UA | Week 1" = one_sample_per_subject(metadata, c("UA"), week = c(1)),
#                 "UA | Week 2" = one_sample_per_subject(metadata, c("UA"), week = c(2)),
#                 "UA | Week 3" = one_sample_per_subject(metadata, c("UA"), week = c(3)),
#                 "UA | Week 4" = one_sample_per_subject(metadata, c("UA"), week = c(4)))

samples <- list("BAL_and_TA | Week 1" = one_sample_per_subject(metadata, c("BAL","TA"), week = c(1)),
                "BAL_and_TA | Week 2" = one_sample_per_subject(metadata, c("BAL","TA"), week = c(2)),
                "BAL_and_TA | Week 3+" = one_sample_per_subject(metadata, c("BAL","TA"), week = c(3,4)),
                "UA | Week 1" = one_sample_per_subject(metadata, c("UA"), week = c(1)),
                "UA | Week 2" = one_sample_per_subject(metadata, c("UA"), week = c(2)),
                "UA | Week 3+" = one_sample_per_subject(metadata, c("UA"), week = c(3,4)))
```

## Plot sample distribution

```{r,fig.height=5,fig.width=8}
plot.df <- as.data.frame(cbind(names(samples),
                               unlist(lapply(samples,length))))
plot.df[,1] <- factor(plot.df[,1],levels=rev(names(samples)))
plot.df[,2] <- as.numeric(as.character(plot.df[,2]))

for(i in 1:nrow(plot.df)){
  metadata.subset <- metadata[match(samples[[i]],metadata$Sample.ID),]
  plot.df[i,3] <- nrow(metadata.subset[metadata.subset$dead_or_alive == "Alive",])
  plot.df[i,4] <- nrow(metadata.subset[metadata.subset$dead_or_alive == "Dead",])
}
colnames(plot.df) <- c("Sample","# of Samples","Alive","Dead")
plot.df <- reshape2::melt(plot.df)
plot.df <- plot.df[plot.df[,2] != "# of Samples",]
p3 <- ggplot()+
  geom_bar(aes(x=!!plot.df[,3],
               y=!!plot.df[,1],
               fill=!!plot.df[,2]),
           color="black",
           stat="identity")+
  labs(x="# of samples",fill = "Outcome")+
  theme_bw()+
  theme(axis.title.y = element_blank(),
        legend.position="bottom")
```
# Plot stats
```{r,fig.height=9,fig.width=9}
ggarrange(p1a,p1b,
          ggarrange(p2,p3,
                    ncol=1,
                    labels=c("B","C")),
          ncol=3,
          widths=c(5,1,5),
          labels=c("A"))
```

# Look at overall differences

## Run MDS

```{r}
mds <- list()

for(i in 1:length(sample_types)){
  metadata.subset <- metadata[metadata$sample_type %in% sample_types[[i]],]
  counts.subset <- counts[filter_taxa[[i]],metadata.subset$Sample.ID]
  
  groups <- factor(metadata.subset$is_immunosuppressed, levels=c("N","Y"))

  design <- model.matrix(~groups)

  colData <- as.data.frame(cbind(groups))
  rownames(colData) <- colnames(counts.subset)
  colnames(colData) <- c("groups")
  
  ddsv <- DESeqDataSetFromMatrix(countData = round(counts.subset),
                                 colData = colData,
                                 design = design)
  idx <- which(colSums(counts(ddsv)) == 0)
  if(length(idx) > 0){ddsv <- ddsv[ , -idx]}
  
  gm_mean = function(x, na.rm=TRUE){ exp(sum(log(x[x > 0]), na.rm=na.rm) / length(x))}
  geoMeans = apply(counts(ddsv), 1, gm_mean)
  
  ddsv <- estimateSizeFactors(ddsv, geoMeans = geoMeans)
  vsdv <- varianceStabilizingTransformation(ddsv)
  vsdv0 <- ifelse(assay(vsdv)<0,0,assay(vsdv))
  
  vegdist <- vegdist(as.data.frame(t(vsdv0)),method="bray")
  cmdScale <- cmdscale(vegdist, k=ifelse(ncol(counts) <= 10,ncol(counts)-1,10))
  vars <- apply(cmdScale, 2, var)
  percentVar <- round(100 * (vars/sum(vars)))
  
  mds[[i]] <- list()
  mds[[i]]$vsdv <- vsdv
  mds[[i]]$vegdist <- vegdist
  mds[[i]]$cmdScale <- cmdScale
  mds[[i]]$vars <- vars
  mds[[i]]$percentVar <- percentVar
}
names(mds) <- names(sample_types)
```

## Use ADONIS to check for diversity metrics


```{r,fig.height=6,fig.width=11}
adonis_categories <- c("weekpostintub","is_immunosuppressed","age_group","bmi_group","hx_COVIDvax","dead_or_alive",
                       "was_trached","had_ecmo","sex")

metadata$weekpostintub <- ifelse(metadata$weekpostintub >=3, "3+",metadata$weekpostintub)
# metadata$bmi_group <- ifelse(metadata$bmi >= 18.9 & metadata$bmi < 25,"Normal","Abnormal")
metadata$bmi_group <- ifelse(metadata$bmi < 25,"<25","25+")
metadata$age_group <- ifelse(metadata$age >= 65,"65+","<65")
```

### Alpha diversity
```{r,fig.height=6,fig.width=11}
alpha.plots.list <- list()
for(i in 1:length(sample_types)){
  metadata.subset <- metadata[metadata$sample_type %in% sample_types[[i]],]
  
  alpha.plots.list[[i]] <- list()
  data_shannon <- diversity(t(counts[,metadata.subset$Sample.ID]), index = "shannon") 

  for(j in 1:length(adonis_categories)){
    groups <- metadata.subset[,adonis_categories[j]]
  
    plot.df <- as.data.frame(cbind(groups,data_shannon))
    colnames(plot.df) <- c("group","shannon")
    plot.df$shannon <- as.numeric(as.character(plot.df$shannon))
    
    kw <- kruskal.test(shannon ~ group, data = plot.df)
    
    if(length(unique(groups)) > 2){
      col <- c("#e49433","#66c2a5","#1a9850")
    }else{
      col <- c("#377eb8","#e4211c")
    }
    
    alpha.plots.list[[i]][[j]] <- ggplot(plot.df,aes(x=group,y=shannon,fill=group))+
      geom_boxplot(outlier.shape = NA)+
      geom_jitter(size=0.9,width = .05)+
      guides(fill="none")+
      #stat_compare_means(method = "kruskal.test")+
      labs(title=paste0(names(sample_types)[i],"\n",adonis_categories[j],"\np = ",round(kw$p.value,3)))+
      theme_bw()+
      theme(plot.title = element_text(size=8),
            axis.title.x = element_blank())+
      scale_fill_manual(values=col)

  }
  egg::ggarrange(plots = alpha.plots.list[[i]],
                 labels=c(LETTERS[[i]],rep("",length(alpha.plots.list[[i]])-1)),
                 nrow=2)
}

```

### Beta diversity

```{r,fig.height=6,fig.width=11}
beta.plots.list <- list()
for(i in 1:length(sample_types)){
  metadata.subset <- metadata[metadata$sample_type %in% sample_types[[i]],]
  
  beta.plots.list[[i]] <- list()
  for(j in 1:length(adonis_categories)){
    groups <- metadata.subset[,adonis_categories[j]]
    
    newResults <- as.data.frame(cbind(mds[[i]]$cmdScale,
                                      as.character(groups)))
    colnames(newResults)[1] <- "PC1"
    colnames(newResults)[2] <- "PC2"
    colnames(newResults)[ncol(newResults)] <- "groups"
  
    centroids <- aggregate(cbind(as.numeric(as.character(PC1)),
                                 as.numeric(as.character(PC2))) ~ groups,
                           data= newResults, mean)
    colnames(centroids) <- c("groups","PC1","PC2")
    newResults <- merge(newResults,centroids,by="groups",suffixes=c("",".centroid"))
  
    for(k in 2:ncol(newResults)){newResults[,k] <- as.numeric(as.character(newResults[,k]))}
    newResults$labels <- rownames(mds[[i]]$cmdScale)
  
    newResults <- as.data.frame(newResults)
  
    data.adonis <- data.frame(colData(mds[[i]]$vsdv))
    data.adonis$groups <- groups
    
    samplepermanova <- adonis2(mds[[i]]$vegdist ~ groups)
    # samplepermanova <- as.data.frame(samplepermanova$aov.tab)
    samplepermanova <- samplepermanova$'Pr(>F)'[1]
    
    if(length(unique(groups)) > 2){
      col <- c("#e49433","#66c2a5","#1a9850")
    }else{
      col <- c("#377eb8","#e4211c")
    }
    
    beta.plots.list[[i]][[j]] <- ggplot() + # Graph PC1 and PC2
      geom_segment(aes(x=!!newResults$PC1.centroid, y=!!newResults$PC2.centroid,
                       xend=!!newResults$PC1, yend=!!newResults$PC2, color=!!newResults$groups),
                   alpha=0.8)+
      geom_point(aes(x=!!newResults$PC1, y=!!newResults$PC2,
                     color=!!newResults$groups),
                 size=2,alpha=0.5)+
      geom_label_repel(aes(x=!!centroids$PC1, y=!!centroids$PC2, 
                           label=!!centroids$groups,color=!!centroids$groups), size=3,alpha=0.8)+
      labs(title=paste0(names(sample_types)[i],"\n",adonis_categories[j],"\nPr(>F) = ",samplepermanova),
           x=paste0("PC1: ",percentVar[1],"% variance"),
           y=paste0("PC2: ",percentVar[2],"% variance"),
           color=names(sample_types)[i])+
      guides(color="none")+
      theme_bw()+
      theme(plot.title = element_text(size=8))+
      scale_color_manual(values=col)

  }
  egg::ggarrange(plots = beta.plots.list[[i]],
                 labels=c(LETTERS[[i]],rep("",length(beta.plots.list[[i]])-1)),
                 nrow=2)
}
```

```{r,fig.height=11,fig.width=9}

pdf(paste0(WORKING.DIR,"/plots/Fig2.pdf"),
    height=11,
    width=9,
    onefile = F)
egg::ggarrange(alpha.plots.list[[1]][[1]],beta.plots.list[[1]][[1]],alpha.plots.list[[1]][[2]],beta.plots.list[[1]][[2]],
               alpha.plots.list[[1]][[3]],beta.plots.list[[1]][[3]],alpha.plots.list[[1]][[4]],beta.plots.list[[1]][[4]],
               alpha.plots.list[[1]][[5]],beta.plots.list[[1]][[5]],alpha.plots.list[[1]][[6]],beta.plots.list[[1]][[6]],
               alpha.plots.list[[1]][[7]],beta.plots.list[[1]][[7]],alpha.plots.list[[1]][[8]],beta.plots.list[[1]][[8]],
               labels=c("A","","B","","C","","D","",
                        "E","","F","","G","","H",""),
               widths=c(1.5,2,1.5,2),
               nrow=4)
dev.off()


pdf(paste0(WORKING.DIR,"/plots/SupFig3.pdf"),
    height=11,
    width=9,
    onefile = F)
egg::ggarrange(alpha.plots.list[[2]][[1]],beta.plots.list[[2]][[1]],alpha.plots.list[[2]][[2]],beta.plots.list[[2]][[2]],
               alpha.plots.list[[2]][[3]],beta.plots.list[[2]][[3]],alpha.plots.list[[2]][[4]],beta.plots.list[[2]][[4]],
               alpha.plots.list[[2]][[5]],beta.plots.list[[2]][[5]],alpha.plots.list[[2]][[6]],beta.plots.list[[2]][[6]],
               alpha.plots.list[[2]][[7]],beta.plots.list[[2]][[7]],alpha.plots.list[[2]][[8]],beta.plots.list[[2]][[8]],
               labels=c("A","","B","","C","","D","",
                        "E","","F","","G","","H",""),
               widths=c(1.5,2,1.5,2),
               nrow=4)
dev.off()
```



# Look at differences individually for each week

## Run MDS

```{r}
mds <- list()

for(i in 1:length(samples)){
  sample_type <- gsub(" [|].*","",names(samples)[i])
  
  metadata.subset <- metadata[metadata$Sample.ID %in% samples[[i]],]
  counts.subset <- counts[filter_taxa[[sample_type]],metadata.subset$Sample.ID]
  
  groups <- factor(metadata.subset$is_immunosuppressed, levels=c("N","Y"))

  design <- model.matrix(~groups)

  colData <- as.data.frame(cbind(groups))
  rownames(colData) <- colnames(counts.subset)
  colnames(colData) <- c("groups")
  
  ddsv <- DESeqDataSetFromMatrix(countData = round(counts.subset),
                                 colData = colData,
                                 design = design)
  idx <- which(colSums(counts(ddsv)) == 0)
  if(length(idx) > 0){ddsv <- ddsv[ , -idx]}
  
  gm_mean = function(x, na.rm=TRUE){ exp(sum(log(x[x > 0]), na.rm=na.rm) / length(x))}
  geoMeans = apply(counts(ddsv), 1, gm_mean)
  
  ddsv <- estimateSizeFactors(ddsv, geoMeans = geoMeans)
  vsdv <- varianceStabilizingTransformation(ddsv)
  vsdv0 <- ifelse(assay(vsdv)<0,0,assay(vsdv))
  
  vegdist <- vegdist(as.data.frame(t(vsdv0)),method="bray")
  cmdScale <- cmdscale(vegdist, k=ifelse(ncol(counts) <= 10,ncol(counts)-1,10))
  vars <- apply(cmdScale, 2, var)
  percentVar <- round(100 * (vars/sum(vars)))
  
  mds[[i]] <- list()
  mds[[i]]$vsdv <- vsdv
  mds[[i]]$vegdist <- vegdist
  mds[[i]]$cmdScale <- cmdScale
  mds[[i]]$vars <- vars
  mds[[i]]$percentVar <- percentVar
}

names(mds) <- names(sample_types)
```

## Use ADONIS to check for diversity metrics

```{r,fig.height=6,fig.width=11}
adonis_categories <- c("is_immunosuppressed","age_group","bmi_group","hx_COVIDvax","dead_or_alive",
                       "was_trached","had_ecmo","sex")
```

### Alpha diversity

```{r,fig.height=6,fig.width=11}
alpha.plots.list <- list()
for(i in 1:length(mds)){
  metadata.subset <- metadata[match(samples[[i]],metadata$Sample.ID),]
  
  alpha.plots.list[[i]] <- list()
  data_shannon <- diversity(t(counts[,metadata.subset$Sample.ID]), index = "shannon") 

  for(j in 1:length(adonis_categories)){
    groups <- metadata.subset[,adonis_categories[j]]
  
    plot.df <- as.data.frame(cbind(groups,data_shannon))
    colnames(plot.df) <- c("group","shannon")
    plot.df$shannon <- as.numeric(as.character(plot.df$shannon))
    
    kw <- kruskal.test(shannon ~ group, data = plot.df)
    
    if(length(unique(groups)) > 2){
      col <- c("#e49433","#66c2a5","#1a9850")
    }else{
      col <- c("#377eb8","#e4211c")
    }
    
    alpha.plots.list[[i]][[j]] <- ggplot(plot.df,aes(x=group,y=shannon,fill=group))+
      geom_boxplot(outlier.shape = NA)+
      geom_jitter(size=0.9,width = .05)+
      guides(fill="none")+
      #stat_compare_means(method = "kruskal.test")+
      labs(title=paste0(names(samples)[i],"\n",adonis_categories[j],"\np = ",round(kw$p.value,2)))+
      theme_bw()+
      theme(plot.title = element_text(size=8),
            axis.title.x = element_blank())+
      scale_fill_manual(values=col)
  }
  egg::ggarrange(plots = alpha.plots.list[[i]],
                 labels=c(LETTERS[[i]],rep("",length(alpha.plots.list[[i]])-1)),
                 nrow=2)
}


```


### Beta diversity
```{r,fig.height=6,fig.width=11}
beta.plots.list <- list()
for(i in 1:length(mds)){
  
  metadata.subset <- metadata[match(samples[[i]],metadata$Sample.ID),]
  
  beta.plots.list[[i]] <- list()
  for(j in 1:length(adonis_categories)){
    groups <- metadata.subset[,adonis_categories[j]]
    
    newResults <- as.data.frame(cbind(mds[[i]]$cmdScale,
                                      as.character(groups)))
    colnames(newResults)[1] <- "PC1"
    colnames(newResults)[2] <- "PC2"
    colnames(newResults)[ncol(newResults)] <- "groups"
  
    centroids <- aggregate(cbind(as.numeric(as.character(PC1)),
                                 as.numeric(as.character(PC2))) ~ groups,
                           data= newResults, mean)
    colnames(centroids) <- c("groups","PC1","PC2")
    newResults <- merge(newResults,centroids,by="groups",suffixes=c("",".centroid"))
  
    for(k in 2:ncol(newResults)){newResults[,k] <- as.numeric(as.character(newResults[,k]))}
    newResults$labels <- rownames(mds[[i]]$cmdScale)
  
    newResults <- as.data.frame(newResults)
  
    data.adonis <- data.frame(colData(mds[[i]]$vsdv))
    data.adonis$groups <- groups
    
    samplepermanova <- adonis2(mds[[i]]$vegdist ~ groups)
    # samplepermanova <- as.data.frame(samplepermanova$aov.tab)
    
    samplepermanova <- samplepermanova$'Pr(>F)'[1]
    if(length(unique(groups)) > 2){
      col <- c("#e49433","#66c2a5","#1a9850")
    }else{
      col <- c("#377eb8","#e4211c")
    }
    
    beta.plots.list[[i]][[j]] <- ggplot() + # Graph PC1 and PC2
      geom_segment(aes(x=!!newResults$PC1.centroid, y=!!newResults$PC2.centroid,
                       xend=!!newResults$PC1, yend=!!newResults$PC2, color=!!newResults$groups),
                   alpha=0.8)+
      geom_point(aes(x=!!newResults$PC1, y=!!newResults$PC2,
                     color=!!newResults$groups),
                 size=2,alpha=0.5)+
      geom_label_repel(aes(x=!!centroids$PC1, y=!!centroids$PC2, 
                           label=!!centroids$groups,color=!!centroids$groups), size=3,alpha=0.8)+
      labs(title=paste0(names(samples)[i],"\n",adonis_categories[j],"\nPr(>F) = ",samplepermanova),
           x=paste0("PC1: ",percentVar[1],"% variance"),
           y=paste0("PC2: ",percentVar[2],"% variance"),
           color=names(samples)[i])+
      guides(color="none")+
      theme_bw()+
      theme(plot.title = element_text(size=8))+
      scale_color_manual(values = col)

  }
  egg::ggarrange(plots = beta.plots.list[[i]],
                 labels=LETTERS[1:length(adonis_categories)],
                 nrow=2)
}
```


```{r,fig.height=11,fig.width=15}
labs <- for(x in 1:21){ifelse((x-1)%%3 != 0,LETTERS[[x]],"")}

labs <- c()
for (x in 1:21) { 
  if((x-1)%%3 == 0) {
    labs <- c(labs,LETTERS[[(x-1)/3+1]])
  }else{
    labs <- c(labs,"")
  }
}

pdf(paste0(WORKING.DIR,"/plots/SupFig4.pdf"),
    height=11,
    width=15,
    onefile = F)
egg::ggarrange(alpha.plots.list[[1]][[1]],alpha.plots.list[[2]][[1]],alpha.plots.list[[3]][[1]],
               alpha.plots.list[[1]][[2]],alpha.plots.list[[2]][[2]],alpha.plots.list[[3]][[2]],
               alpha.plots.list[[1]][[3]],alpha.plots.list[[2]][[3]],alpha.plots.list[[3]][[3]],
               alpha.plots.list[[1]][[4]],alpha.plots.list[[2]][[4]],alpha.plots.list[[3]][[4]],
               alpha.plots.list[[1]][[5]],alpha.plots.list[[2]][[5]],alpha.plots.list[[3]][[5]],
               alpha.plots.list[[1]][[6]],alpha.plots.list[[2]][[6]],alpha.plots.list[[3]][[6]],   
               alpha.plots.list[[1]][[7]],alpha.plots.list[[2]][[7]],alpha.plots.list[[3]][[7]],
               labels=labs,
               ncol=6)
dev.off()

pdf(paste0(WORKING.DIR,"/plots/SupFig5.pdf"),
    height=11,
    width=15,
    onefile = F)
egg::ggarrange(alpha.plots.list[[4]][[1]],alpha.plots.list[[5]][[1]],alpha.plots.list[[6]][[1]],
               alpha.plots.list[[4]][[2]],alpha.plots.list[[5]][[2]],alpha.plots.list[[6]][[2]],
               alpha.plots.list[[4]][[3]],alpha.plots.list[[5]][[3]],alpha.plots.list[[6]][[3]],
               alpha.plots.list[[4]][[4]],alpha.plots.list[[5]][[4]],alpha.plots.list[[6]][[4]],
               alpha.plots.list[[4]][[5]],alpha.plots.list[[5]][[5]],alpha.plots.list[[6]][[5]],
               alpha.plots.list[[4]][[6]],alpha.plots.list[[5]][[6]],alpha.plots.list[[6]][[6]],   
               alpha.plots.list[[4]][[7]],alpha.plots.list[[5]][[7]],alpha.plots.list[[6]][[7]],
               labels=labs,
               ncol=6)
dev.off()
```    



```{r,fig.height=11,fig.width=15}
labs <- for(x in 1:21){ifelse((x-1)%%3 != 0,LETTERS[[x]],"")}

labs <- c()
for (x in 1:21) { 
  if((x-1)%%3 == 0) {
    labs <- c(labs,LETTERS[[(x-1)/3+1]])
  }else{
    labs <- c(labs,"")
  }
}

pdf(paste0(WORKING.DIR,"/plots/SupFig6.pdf"),
    height=11,
    width=15,
    onefile = F)
egg::ggarrange(beta.plots.list[[1]][[1]],beta.plots.list[[2]][[1]],beta.plots.list[[3]][[1]],
               beta.plots.list[[1]][[2]],beta.plots.list[[2]][[2]],beta.plots.list[[3]][[2]],
               beta.plots.list[[1]][[3]],beta.plots.list[[2]][[3]],beta.plots.list[[3]][[3]],
               beta.plots.list[[1]][[4]],beta.plots.list[[2]][[4]],beta.plots.list[[3]][[4]],
               beta.plots.list[[1]][[5]],beta.plots.list[[2]][[5]],beta.plots.list[[3]][[5]],
               beta.plots.list[[1]][[6]],beta.plots.list[[2]][[6]],beta.plots.list[[3]][[6]],   
               beta.plots.list[[1]][[7]],beta.plots.list[[2]][[7]],beta.plots.list[[3]][[7]],
               labels=labs,
               ncol=6)
dev.off()
        

pdf(paste0(WORKING.DIR,"/plots/SupFig7.pdf"),
    height=11,
    width=15,
    onefile = F)
egg::ggarrange(beta.plots.list[[4]][[1]],beta.plots.list[[5]][[1]],beta.plots.list[[6]][[1]],
               beta.plots.list[[4]][[2]],beta.plots.list[[5]][[2]],beta.plots.list[[6]][[2]],
               beta.plots.list[[4]][[3]],beta.plots.list[[5]][[3]],beta.plots.list[[6]][[3]],
               beta.plots.list[[4]][[4]],beta.plots.list[[5]][[4]],beta.plots.list[[6]][[4]],
               beta.plots.list[[4]][[5]],beta.plots.list[[5]][[5]],beta.plots.list[[6]][[5]],
               beta.plots.list[[4]][[6]],beta.plots.list[[5]][[6]],beta.plots.list[[6]][[6]],   
               beta.plots.list[[4]][[7]],beta.plots.list[[5]][[7]],beta.plots.list[[6]][[7]],
               labels=labs,
               ncol=6)
dev.off()
        
```


# Look at differences individually for each week specifically in immunosuppressed

## Run MDS

```{r}
mds <- list()

for(i in 1:length(samples)){
  sample_type <- gsub(" [|].*","",names(samples)[i])
  
  metadata.subset <- metadata[metadata$Sample.ID %in% samples[[i]],]
  metadata.subset <- metadata.subset[metadata.subset$is_immunosuppressed == "N",]
  
  counts.subset <- counts[filter_taxa[[sample_type]],metadata.subset$Sample.ID]
  
  # groups <- factor(metadata.subset$is_immunosuppressed, levels=c("N","Y"))
  groups <- factor(metadata.subset$dead_or_alive, levels=c("Alive","Dead"))

  design <- model.matrix(~groups)

  colData <- as.data.frame(cbind(groups))
  rownames(colData) <- colnames(counts.subset)
  colnames(colData) <- c("groups")
  
  ddsv <- DESeqDataSetFromMatrix(countData = round(counts.subset),
                                 colData = colData,
                                 design = design)
  idx <- which(colSums(counts(ddsv)) == 0)
  if(length(idx) > 0){ddsv <- ddsv[ , -idx]}
  
  gm_mean = function(x, na.rm=TRUE){ exp(sum(log(x[x > 0]), na.rm=na.rm) / length(x))}
  geoMeans = apply(counts(ddsv), 1, gm_mean)
  
  ddsv <- estimateSizeFactors(ddsv, geoMeans = geoMeans)
  vsdv <- varianceStabilizingTransformation(ddsv)
  vsdv0 <- ifelse(assay(vsdv)<0,0,assay(vsdv))
  
  vegdist <- vegdist(as.data.frame(t(vsdv0)),method="bray")
  cmdScale <- cmdscale(vegdist, k=ifelse(ncol(counts.subset) <= 10,ncol(counts.subset)-1,10))
  vars <- apply(cmdScale, 2, var)
  percentVar <- round(100 * (vars/sum(vars)))
  
  mds[[i]] <- list()
  mds[[i]]$vsdv <- vsdv
  mds[[i]]$vegdist <- vegdist
  mds[[i]]$cmdScale <- cmdScale
  mds[[i]]$vars <- vars
  mds[[i]]$percentVar <- percentVar
}

names(mds) <- names(sample_types)
```


## Use ADONIS to check for diversity metrics


```{r,fig.height=6,fig.width=11}
adonis_categories <- c("age_group","bmi_group","hx_COVIDvax","dead_or_alive",
                       "was_trached","had_ecmo","sex")
```


### Alpha diversity

```{r,fig.height=6,fig.width=11}
alpha.plots.list <- list()
for(i in 1:length(mds)){
  metadata.subset <- metadata[match(samples[[i]],metadata$Sample.ID),]
  metadata.subset <- metadata.subset[metadata.subset$is_immunosuppressed == "N",]

  alpha.plots.list[[i]] <- list()
  data_shannon <- diversity(t(counts[,metadata.subset$Sample.ID]), index = "shannon") 

  for(j in 1:length(adonis_categories)){
    groups <- metadata.subset[,adonis_categories[j]]
  
    plot.df <- as.data.frame(cbind(groups,data_shannon))
    colnames(plot.df) <- c("group","shannon")
    plot.df$shannon <- as.numeric(as.character(plot.df$shannon))
    
    kw <- kruskal.test(shannon ~ group, data = plot.df)
    
    alpha.plots.list[[i]][[j]] <- ggplot(plot.df,aes(x=group,y=shannon,fill=group))+
      geom_boxplot(outlier.shape = NA)+
      geom_jitter(width = .05)+
      guides(fill="none")+
      #stat_compare_means(method = "kruskal.test")+
      labs(title=paste0(names(samples)[i],"\n",adonis_categories[j],"\np = ",round(kw$p.value,2)))+
      theme_bw()+
      theme(plot.title = element_text(size=8),
            axis.title.x = element_blank())+
      scale_fill_brewer(palette="Set1",direction = -1)

  }
  egg::ggarrange(plots = alpha.plots.list[[i]],
                 labels=c(LETTERS[[i]],rep("",length(alpha.plots.list[[i]])-1)),
                 nrow=2)
}


```

### Beta diversity
```{r,fig.height=6,fig.width=11}
beta.plots.list <- list()
for(i in 1:length(mds)){
  metadata.subset <- metadata[match(samples[[i]],metadata$Sample.ID),]
  metadata.subset <- metadata.subset[metadata.subset$is_immunosuppressed == "N",]

  beta.plots.list[[i]] <- list()
  for(j in 1:length(adonis_categories)){
    groups <- metadata.subset[,adonis_categories[j]]
    
    newResults <- as.data.frame(cbind(mds[[i]]$cmdScale,
                                      as.character(groups)))
    colnames(newResults)[1] <- "PC1"
    colnames(newResults)[2] <- "PC2"
    colnames(newResults)[ncol(newResults)] <- "groups"
  
    centroids <- aggregate(cbind(as.numeric(as.character(PC1)),
                                 as.numeric(as.character(PC2))) ~ groups,
                           data= newResults, mean)
    colnames(centroids) <- c("groups","PC1","PC2")
    newResults <- merge(newResults,centroids,by="groups",suffixes=c("",".centroid"))
  
    for(k in 2:ncol(newResults)){newResults[,k] <- as.numeric(as.character(newResults[,k]))}
    newResults$labels <- rownames(mds[[i]]$cmdScale)
  
    newResults <- as.data.frame(newResults)
  
    data.adonis <- data.frame(colData(mds[[i]]$vsdv))
    data.adonis$groups <- groups
    
    samplepermanova <- adonis2(mds[[i]]$vegdist ~ groups)
    # samplepermanova <- as.data.frame(samplepermanova$aov.tab)
    samplepermanova <- samplepermanova$'Pr(>F)'[1]

    beta.plots.list[[i]][[j]] <- ggplot() + # Graph PC1 and PC2
      geom_segment(aes(x=!!newResults$PC1.centroid, y=!!newResults$PC2.centroid,
                       xend=!!newResults$PC1, yend=!!newResults$PC2, color=!!newResults$groups),
                   alpha=0.8)+
      geom_point(aes(x=!!newResults$PC1, y=!!newResults$PC2,
                     color=!!newResults$groups),
                 size=2,alpha=0.5)+
      geom_label_repel(aes(x=!!centroids$PC1, y=!!centroids$PC2, 
                           label=!!centroids$groups,color=!!centroids$groups), size=3,alpha=0.8)+
      labs(title=paste0(names(samples)[i],"\n",adonis_categories[j],"\nPr(>F) = ",samplepermanova),
           x=paste0("PC1: ",percentVar[1],"% variance"),
           y=paste0("PC2: ",percentVar[2],"% variance"),
           color=names(samples)[i])+
      guides(color="none")+
      theme_bw()+
      theme(plot.title = element_text(size=8))+
      scale_color_brewer(palette="Set1",direction = -1)
  }
  egg::ggarrange(plots = beta.plots.list[[i]],
                 labels=LETTERS[1:length(adonis_categories)],
                 nrow=2)
}
```

```{r,fig.height=11,fig.width=15}
labs <- for(x in 1:21){ifelse((x-1)%%3 != 0,LETTERS[[x]],"")}

labs <- c()
for (x in 1:21) { 
  if((x-1)%%3 == 0) {
    labs <- c(labs,LETTERS[[(x-1)/3+1]])
  }else{
    labs <- c(labs,"")
  }
}
# 
# pdf(paste0(WORKING.DIR,"/plots/SupFig4.pdf"),
#     height=11,
#     width=15,
#     onefile = F)
egg::ggarrange(alpha.plots.list[[1]][[1]],alpha.plots.list[[2]][[1]],alpha.plots.list[[3]][[1]],
               alpha.plots.list[[1]][[2]],alpha.plots.list[[2]][[2]],alpha.plots.list[[3]][[2]],
               alpha.plots.list[[1]][[3]],alpha.plots.list[[2]][[3]],alpha.plots.list[[3]][[3]],
               alpha.plots.list[[1]][[4]],alpha.plots.list[[2]][[4]],alpha.plots.list[[3]][[4]],
               alpha.plots.list[[1]][[5]],alpha.plots.list[[2]][[5]],alpha.plots.list[[3]][[5]],
               alpha.plots.list[[1]][[6]],alpha.plots.list[[2]][[6]],alpha.plots.list[[3]][[6]],   
               alpha.plots.list[[1]][[7]],alpha.plots.list[[2]][[7]],alpha.plots.list[[3]][[7]],
               labels=labs,
               ncol=6)
# dev.off()

pdf(paste0(WORKING.DIR,"/plots/SupFig5.pdf"),
    height=11,
    width=15,
    onefile = F)
egg::ggarrange(alpha.plots.list[[4]][[1]],alpha.plots.list[[5]][[2]],alpha.plots.list[[6]][[2]],
               alpha.plots.list[[4]][[2]],alpha.plots.list[[5]][[2]],alpha.plots.list[[6]][[2]],
               alpha.plots.list[[4]][[3]],alpha.plots.list[[5]][[3]],alpha.plots.list[[6]][[3]],
               alpha.plots.list[[4]][[4]],alpha.plots.list[[5]][[4]],alpha.plots.list[[6]][[4]],
               alpha.plots.list[[4]][[5]],alpha.plots.list[[5]][[5]],alpha.plots.list[[6]][[5]],
               alpha.plots.list[[4]][[6]],alpha.plots.list[[5]][[6]],alpha.plots.list[[6]][[6]],   
               alpha.plots.list[[4]][[7]],alpha.plots.list[[5]][[7]],alpha.plots.list[[6]][[7]],
               labels=labs,
               ncol=6)
dev.off()
```    


```{r,fig.height=11,fig.width=15}
labs <- for(x in 1:21){ifelse((x-1)%%3 != 0,LETTERS[[x]],"")}

labs <- c()
for (x in 1:21) { 
  if((x-1)%%3 == 0) {
    labs <- c(labs,LETTERS[[(x-1)/3+1]])
  }else{
    labs <- c(labs,"")
  }
}

pdf(paste0(WORKING.DIR,"/plots/SupFig6.pdf"),
    height=11,
    width=15,
    onefile = F)
egg::ggarrange(beta.plots.list[[1]][[1]],beta.plots.list[[2]][[1]],beta.plots.list[[3]][[1]],
               beta.plots.list[[1]][[2]],beta.plots.list[[2]][[2]],beta.plots.list[[3]][[2]],
               beta.plots.list[[1]][[3]],beta.plots.list[[2]][[3]],beta.plots.list[[3]][[3]],
               beta.plots.list[[1]][[4]],beta.plots.list[[2]][[4]],beta.plots.list[[3]][[4]],
               beta.plots.list[[1]][[5]],beta.plots.list[[2]][[5]],beta.plots.list[[3]][[5]],
               beta.plots.list[[1]][[6]],beta.plots.list[[2]][[6]],beta.plots.list[[3]][[6]],   
               beta.plots.list[[1]][[7]],beta.plots.list[[2]][[7]],beta.plots.list[[3]][[7]],
               labels=labs,
               ncol=6)
dev.off()
        

pdf(paste0(WORKING.DIR,"/plots/SupFig7.pdf"),
    height=11,
    width=15,
    onefile = F)
egg::ggarrange(beta.plots.list[[4]][[1]],beta.plots.list[[5]][[1]],beta.plots.list[[6]][[1]],
               beta.plots.list[[4]][[2]],beta.plots.list[[5]][[2]],beta.plots.list[[6]][[2]],
               beta.plots.list[[4]][[3]],beta.plots.list[[5]][[3]],beta.plots.list[[6]][[3]],
               beta.plots.list[[4]][[4]],beta.plots.list[[5]][[4]],beta.plots.list[[6]][[4]],
               beta.plots.list[[4]][[5]],beta.plots.list[[5]][[5]],beta.plots.list[[6]][[5]],
               beta.plots.list[[4]][[6]],beta.plots.list[[5]][[6]],beta.plots.list[[6]][[6]],   
               beta.plots.list[[4]][[7]],beta.plots.list[[5]][[7]],beta.plots.list[[6]][[7]],
               labels=labs,
               ncol=6)
dev.off()
        
```
