---
title: "DE analysis"
output: html_document
---

# Preanalysis

## Set inputs
```{r}
# library(ANCOMBC)
library(ALDEx2)
library(circlize)
library(ComplexHeatmap)
library(DESeq2)
# library(dplyr)
# library(egg)
library(edgeR)
library(ggplot2)
library(ggrepel)
# library(limma)
# library(matrixStats)
library(phyloseq)
library(rlist)
library(UpSetR)
# library(vegan)

WORKING.DIR="Z:/mchung/leopoldo_segal_07/"
```

## Load upstream object
```{r}
load(paste0(WORKING.DIR,"/upstream.R"))
```

```{r}
counts <- counts$Bracken
rownames(counts) <- gsub("^[[]Candida[]] glabrata$","Nakaseomyces glabratus",rownames(counts))

annotation <- annotation$Bracken
annotation$taxa <- gsub("^[[]Candida[]] glabrata$","Nakaseomyces glabratus",annotation$taxa)
relab <- as.data.frame(apply(counts,2,function(x){x/sum(x)*100}))
# 
# write.table(counts,
#             paste0(WORKING.DIR,"/counts.taxa.bracken.tsv"),
#             row.names = T,
#             col.names = T,
#             quote = F,
#             sep = "\t")
# 
# write.table(metadata,
#             paste0(WORKING.DIR,"/metadata.tsv"),
#             row.names = F,
#             col.names = T,
#             quote = F,
#             sep = "\t")
```

## Filter low prevalence (<1%) and low relative abundance (<0.1%) taxa
```{r}
sample_types <- list("LA" = c("BAL","TA"),
                     "UA" = c("UA"))

filter_taxa <- list()
for(i in 1:length(sample_types)){
  counts.subset <- counts[,metadata$Sample.ID[metadata$sample_type %in% sample_types[[i]]]]
  
  relab.subset <- apply(counts.subset,2,function(x){x/sum(x)*100})
  relab.taxa <- rownames(relab.subset)[rowMeans(relab.subset) >= 0.1]
  
  prev <- rowSums(as.data.frame(ifelse(counts.subset > 0,1,0)))/ncol(counts.subset)*100
  prev.taxa <- rownames(counts.subset)[prev >= 1.0]

  filter_taxa[[i]] <- intersect(relab.taxa,prev.taxa)
}
names(filter_taxa) <- names(sample_types)
```

## Set longitudinal sample list

```{r,fig.height=4,fig.width=11}
extract_longitudinal_samples <- function(metadata,sample_types){
  metadata <- metadata[metadata$sample_type %in% sample_types,]
  samples <- as.data.frame(table(metadata$subject_id))
  samples <- samples[samples[,2] >= 3,1]
  
  metadata <- metadata[order(metadata$subject_id,metadata$intub_to_sample_days),]
  metadata <- metadata[metadata$subject_id %in% samples,]

  return(metadata$Sample.ID)
}

samples_longitudinal <- list("LA" = extract_longitudinal_samples(metadata, c("BAL","TA")),
                             "UA" = extract_longitudinal_samples(metadata, c("UA")))
```

## Set single sample list

```{r}
one_sample_per_subject <- function(metadata,sample_types, week){
  metadata <- metadata[metadata$sample_type %in% sample_types,]
  metadata <- metadata[metadata$weekpostintub %in% week,]
  metadata <- metadata[!duplicated(metadata$subject_id),]
  return(metadata$Sample.ID)
}

# samples <- list("BAL_and_TA | Week 1" = one_sample_per_subject(metadata, c("BAL","TA"), week = c(1)),
#                 "BAL_and_TA | Week 2" = one_sample_per_subject(metadata, c("BAL","TA"), week = c(2)),
#                 "BAL_and_TA | Week 3" = one_sample_per_subject(metadata, c("BAL","TA"), week = c(3)),
#                 "BAL_and_TA | Week 4" = one_sample_per_subject(metadata, c("BAL","TA"), week = c(4)),
#                 "UA | Week 1" = one_sample_per_subject(metadata, c("UA"), week = c(1)),
#                 "UA | Week 2" = one_sample_per_subject(metadata, c("UA"), week = c(2)),
#                 "UA | Week 3" = one_sample_per_subject(metadata, c("UA"), week = c(3)),
#                 "UA | Week 4" = one_sample_per_subject(metadata, c("UA"), week = c(4)))

samples <- list("LA | Week 1" = one_sample_per_subject(metadata, c("BAL","TA"), week = c(1)),
                "LA | Week 2" = one_sample_per_subject(metadata, c("BAL","TA"), week = c(2)),
                "LA | Week 3+" = one_sample_per_subject(metadata, c("BAL","TA"), week = c(3,4)),
                "UA | Week 1" = one_sample_per_subject(metadata, c("UA"), week = c(1)),
                "UA | Week 2" = one_sample_per_subject(metadata, c("UA"), week = c(2)),
                "UA | Week 3+" = one_sample_per_subject(metadata, c("UA"), week = c(3,4)))
```

## Plot sample distribution

```{r,fig.height=4,fig.width=9}
plot.df <- as.data.frame(cbind(names(samples),
                               unlist(lapply(samples,length))))
plot.df[,1] <- factor(plot.df[,1],levels=rev(names(samples)))
plot.df[,2] <- as.numeric(as.character(plot.df[,2]))

for(i in 1:nrow(plot.df)){
  metadata.subset <- metadata[match(samples[[i]],metadata$Sample.ID),]
  plot.df[i,3] <- nrow(metadata.subset[metadata.subset$dead_or_alive == "Alive",])
  plot.df[i,4] <- nrow(metadata.subset[metadata.subset$dead_or_alive == "Dead",])
}
colnames(plot.df) <- c("Sample","# of Samples","Alive","Dead")
plot.df <- reshape2::melt(plot.df)
plot.df <- plot.df[plot.df[,2] != "# of Samples",]
p1 <- ggplot()+
  geom_bar(aes(x=!!plot.df[,3],
               y=!!plot.df[,1],
               fill=!!plot.df[,2]),
           color="black",
           stat="identity")+
  labs(title="# of Samples for edgeR",x="# of samples",fill = "Outcome")+
  theme_bw()+
  theme(axis.title.y = element_blank())

plot.df <- as.data.frame(cbind(names(samples),
                               unlist(lapply(samples,length))))
plot.df[,1] <- factor(plot.df[,1],levels=rev(names(samples)))
plot.df[,2] <- as.numeric(as.character(plot.df[,2]))

for(i in 1:nrow(plot.df)){
  metadata.subset <- metadata[match(samples[[i]],metadata$Sample.ID),]
  plot.df[i,3] <- nrow(metadata.subset[metadata.subset$dead_or_alive == "Alive",])
  plot.df[i,4] <- nrow(metadata.subset[metadata.subset$dead_or_alive == "Dead",])
}

p2 <- ggplot()+
  geom_bar(aes(x=plot.df[,3]/plot.df[,2] * 100,
               y=plot.df[,1]),
           stat="identity",
           color="black",
           fill = "red")+
  labs(title="%tage of Dead",x="% of dead samples")+
  theme_bw()+
  theme(axis.title.y = element_blank())
egg::ggarrange(p1,p2,nrow=1,labels=LETTERS[1:2],
               widths = c(5,2))
```

## Plot counts distribution
```{r,fig.height=5,fig.width=8}
col1 <- c()
col2 <- c()
col3 <- c()
col4 <- c()
for(i in 1:length(samples)){
  metadata.subset <- metadata[match(samples[[i]],metadata$Sample.ID),]
  col1 <- c(col1,samples[[i]])
  col2 <- c(col2,rep(names(samples)[i],length(samples[[i]])))
  col3 <- c(col3,colSums(counts[,samples[[i]]]))
  col4 <- c(col4,metadata.subset$dead_or_alive[match(samples[[i]],metadata.subset$Sample.ID)])
}
plot.df <- as.data.frame(cbind(col1,col2,col3,col4))
plot.df[,2] <- factor(plot.df[,2],levels=rev(names(samples)))
plot.df[,3] <- as.numeric(as.character(plot.df[,3]))
plot.df[,5] <- gsub(" [|].*","",plot.df[,2])
colnames(plot.df) <- c("sample","group","counts","dead_or_alive","sample_type")

ggplot(plot.df,aes(counts,group,fill=dead_or_alive))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(alpha=0.2)+
  facet_wrap(sample_type ~ .,ncol = 1,scales="free")+
  labs(title="Taxa Counts",fill="Outcome")+
  theme_bw()+
  theme(axis.title.y = element_blank())
```


# KO DE analysis

## Load counts
```{r}
load("X:/mchung/leopoldo_segal_07/counts_conqur_KO.R")

counts.raw <- counts.ko$pathway_metabolism
rownames(counts.raw) <- gsub(".*: ","",rownames(counts.raw))
counts.ko<- counts.raw[rowSums(counts.raw) >= 1000,]
```


## Run edgeR

### Comparing Healthy v. Immunosuppressed

```{r}
edgeR.output <- list()
for(i in 1:length(samples)){
  sample_type <- gsub(" [|].*","",names(samples)[i])

  counts.subset <- counts.ko[,samples[[i]]]
  counts.subset <- counts.subset[,colSums(counts.subset) >= 2000]
  metadata.subset <- metadata[match(colnames(counts.subset),metadata$Sample.ID),]
  
  imsupp <- factor(metadata.subset$is_immunosuppressed, levels=c("N","Y"))
  #weekpostintub <- ifelse(metadata.subset$weekpostintub >= 4,3,metadata.subset$weekpostintub)
  # run <- metadata.subset$run
  
  design <- model.matrix(~imsupp)
  # design <- model.matrix(~imsupp + run)
  
  y <- DGEList(counts=counts.subset,
               group=imsupp)
  keep <- filterByExpr(y)  
  y <- y[keep, , keep.lib.sizes = F]
  y <- calcNormFactors(y, method="TMM")
  y <- estimateDisp(y,design, robust=TRUE)
  fit <- glmQLFit(y,design,robust=TRUE)
    
  # qlf <- glmQLFTest(fit,coef=2:length(unique(groups)))
  qlf <- glmQLFTest(fit,coef=2)
  de <- topTags(qlf,n=Inf)$table
  edgeR.output[[i]] <- de
  
  print(length(rownames(edgeR.output[[i]])[edgeR.output[[i]]$FDR < 0.05 & !is.na(edgeR.output[[i]]$FDR)]))
}
names(edgeR.output) <- names(samples)
edgeR.imsupp <- edgeR.output
```


### Comparing Alive v. Dead

```{r}
edgeR.output <- list()
for(i in 1:length(samples)){
  sample_type <- gsub(" [|].*","",names(samples)[i])

  counts.subset <- counts.ko[,samples[[i]]]
  counts.subset <- counts.subset[,colSums(counts.subset) >= 2000]
  metadata.subset <- metadata[match(colnames(counts.subset),metadata$Sample.ID),]
  
  doa <- factor(metadata.subset$dead_or_alive, levels=c("Alive","Dead"))
  #weekpostintub <- ifelse(metadata.subset$weekpostintub >= 4,3,metadata.subset$weekpostintub)
  # run <- metadata.subset$run
  
  design <- model.matrix(~doa)
  # design <- model.matrix(~imsupp + run)
  
  y <- DGEList(counts=counts.subset,
               group=doa)
  keep <- filterByExpr(y)  
  y <- y[keep, , keep.lib.sizes = F]
  y <- calcNormFactors(y, method="TMM")
  y <- estimateDisp(y,design, robust=TRUE)
  fit <- glmQLFit(y,design,robust=TRUE)
    
  # qlf <- glmQLFTest(fit,coef=2:length(unique(groups)))
  qlf <- glmQLFTest(fit,coef=2)
  de <- topTags(qlf,n=Inf)$table
  edgeR.output[[i]] <- de
  
  print(length(rownames(edgeR.output[[i]])[edgeR.output[[i]]$FDR < 0.05 & !is.na(edgeR.output[[i]]$FDR)]))
}
names(edgeR.output) <- names(samples)
edgeR.outcome <- edgeR.output
```

## Plot comparison dotplots

### Compare DE results between Dead v. Alive results as dotplot
```{r,fig.height=5,fig.width=10}
edgeR.output <- edgeR.outcome

ko.plots <- list()
for(h in 1:1){
  sample_sets <- grep(names(sample_types)[h],names(samples))
  
  taxa <- c()
  logFC <- c()
  for(i in sample_sets){
    taxa <- c(taxa,rownames(edgeR.output[[i]])[edgeR.output[[i]]$FDR < 0.05])
    logFC <- c(logFC,edgeR.output[[i]]$logFC[edgeR.output[[i]]$FDR < 0.05])
  }
  score <- ifelse(logFC > 0,1,-1)
  taxa_stats <- as.data.frame(cbind(aggregate(score, by=list(Category=taxa), FUN=sum),
                                    aggregate(logFC, by=list(Category=taxa), FUN=mean)))
  taxa_stats <- taxa_stats[,c(1,2,4)]
  top_taxa <- taxa_stats[order(-abs(taxa_stats[,2]),-abs(taxa_stats[,3])),1] 
  
  taxa <- list()
  for(i in sample_sets){
    taxa[[i]] <- rownames(edgeR.output[[i]])[edgeR.output[[i]]$FDR < 0.05 & !is.na(edgeR.output[[i]]$FDR)]
  }
  taxa <- unique(unlist(taxa))
  taxa <- rev(top_taxa[top_taxa %in% taxa][1:50])
  taxa <- taxa[!is.na(taxa)]
  
  logFC.df <- as.data.frame(matrix(nrow=length(taxa),
                                   ncol=length(sample_sets)))
  rownames(logFC.df) <- taxa
  colnames(logFC.df) <- paste0("Week ",1:length(sample_sets))
  logCPM.df <- logFC.df
  sig.df <- logFC.df
  for(i in 1:length(sample_sets)){
    logFC.df[,i] <- edgeR.output[[sample_sets[i]]]$logFC[match(rownames(logFC.df),rownames(edgeR.output[[sample_sets[i]]]))]
    logCPM.df[,i] <- edgeR.output[[sample_sets[i]]]$logCPM[match(rownames(logCPM.df),rownames(edgeR.output[[sample_sets[i]]]))]
    sig.df[,i] <- ifelse(rownames(logFC.df) %in% rownames(edgeR.output[[sample_sets[i]]])[edgeR.output[[sample_sets[i]]]$FDR < 0.05 & !is.na(edgeR.output[[sample_sets[i]]]$FDR)],T,F)
  }
  
  plot.df <- reshape2::melt(as.matrix(logFC.df))
  colnames(plot.df) <- c("taxa","week","logFC")
  # plot.df$taxa <- factor(plot.df$taxa,levels=rownames(logCPM.df)[order(rowMeans(logCPM.df))])
  plot.df$logCPM <- reshape2::melt(as.matrix(logCPM.df))[,3]
  plot.df$sig <- ifelse(reshape2::melt(as.matrix(sig.df))[,3] == T, plot.df$logFC,NA)
  plot.df$kingdom <- annotation$superkingdom[match(plot.df$taxa,annotation$taxa)]
  plot.df$color <- ifelse(is.na(plot.df$sig),NA,"black")
  plot.df$alpha <- as.numeric(as.character(ifelse(is.na(plot.df$sig),0.2,1)))
  
  plot.df$logCPM <- as.numeric(ifelse(is.na(plot.df$logCPM),0,plot.df$logCPM))
  
  ko.plots[[h]] <- ggplot(plot.df,aes(x=logFC,y=taxa,fill=sig,color=color,alpha=alpha,size=logCPM))+
    geom_segment(aes(xend=0),alpha=0.4,color="black",size=0.7)+
    geom_point(pch=21,stat="identity")+
    geom_vline(aes(xintercept = 0),
               lty="dashed")+
    facet_grid(.~week, scales="free_y",space = "free")+
    scale_color_manual(values = c("black",NA))+
    scale_size_area(limits=c(0,20))+
    scale_fill_gradientn(limits = c(-3,3),colors=c("blue","white","red"),breaks=c(-2,0,2),oob = scales::squish)+
    labs(title=names(sample_types)[h],fill="logFC")+
    guides(alpha="none",color="none")+
    theme_bw()+
    theme(axis.title.y = element_blank(),
          strip.background.y = element_blank(),
          strip.text.y.right = element_text(hjust=0,angle = 0),
          legend.position = "bottom")

  plot(ko.plots[[h]])  
}
```


### Compare DE results between Healthy v. Immunosuppressed results as dotplot
```{r,fig.height=5,fig.width=10}
edgeR.output <- edgeR.imsupp

for(h in 1:1){
  sample_sets <- grep(names(sample_types)[h],names(samples))
  
  taxa <- c()
  logFC <- c()
  for(i in sample_sets){
    taxa <- c(taxa,rownames(edgeR.output[[i]])[edgeR.output[[i]]$FDR < 0.05])
    logFC <- c(logFC,edgeR.output[[i]]$logFC[edgeR.output[[i]]$FDR < 0.05])
  }
  score <- ifelse(logFC > 0,1,-1)
  taxa_stats <- as.data.frame(cbind(aggregate(score, by=list(Category=taxa), FUN=sum),
                                    aggregate(logFC, by=list(Category=taxa), FUN=mean)))
  taxa_stats <- taxa_stats[,c(1,2,4)]
  top_taxa <- taxa_stats[order(-abs(taxa_stats[,2]),-abs(taxa_stats[,3])),1] 
  
  taxa <- list()
  for(i in sample_sets){
    taxa[[i]] <- rownames(edgeR.output[[i]])[edgeR.output[[i]]$FDR < 0.05 & !is.na(edgeR.output[[i]]$FDR)]
  }
  taxa <- unique(unlist(taxa))
  taxa <- rev(top_taxa[top_taxa %in% taxa][1:50])
  taxa <- taxa[!is.na(taxa)]
  
  logFC.df <- as.data.frame(matrix(nrow=length(taxa),
                                   ncol=length(sample_sets)))
  rownames(logFC.df) <- taxa
  colnames(logFC.df) <- paste0("Week ",1:length(sample_sets))
  logCPM.df <- logFC.df
  sig.df <- logFC.df
  for(i in 1:length(sample_sets)){
    logFC.df[,i] <- edgeR.output[[sample_sets[i]]]$logFC[match(rownames(logFC.df),rownames(edgeR.output[[sample_sets[i]]]))]
    logCPM.df[,i] <- edgeR.output[[sample_sets[i]]]$logCPM[match(rownames(logCPM.df),rownames(edgeR.output[[sample_sets[i]]]))]
    sig.df[,i] <- ifelse(rownames(logFC.df) %in% rownames(edgeR.output[[sample_sets[i]]])[edgeR.output[[sample_sets[i]]]$FDR < 0.05 & !is.na(edgeR.output[[sample_sets[i]]]$FDR)],T,F)
  }
  
  plot.df <- reshape2::melt(as.matrix(logFC.df))
  colnames(plot.df) <- c("taxa","week","logFC")
  # plot.df$taxa <- factor(plot.df$taxa,levels=rownames(logCPM.df)[order(rowMeans(logCPM.df))])
  plot.df$logCPM <- reshape2::melt(as.matrix(logCPM.df))[,3]
  plot.df$sig <- ifelse(reshape2::melt(as.matrix(sig.df))[,3] == T, plot.df$logFC,NA)
  plot.df$kingdom <- annotation$superkingdom[match(plot.df$taxa,annotation$taxa)]
  plot.df$color <- ifelse(is.na(plot.df$sig),NA,"black")
  plot.df$alpha <- as.numeric(as.character(ifelse(is.na(plot.df$sig),0.2,1)))
  
  plot.df$logCPM <- as.numeric(ifelse(is.na(plot.df$logCPM),0,plot.df$logCPM))
  
  ko.plots[[2]] <- ggplot(plot.df,aes(x=logFC,y=taxa,fill=sig,color=color,alpha=alpha,size=logCPM))+
    geom_segment(aes(xend=0),alpha=0.4,color="black",size=0.7)+
    geom_point(pch=21,stat="identity")+
    geom_vline(aes(xintercept = 0),
               lty="dashed")+
    facet_grid(.~week, scales="free_y",space = "free")+
    scale_color_manual(values = c("black",NA))+
    scale_size_area(limits=c(0,20))+
    scale_fill_gradientn(limits = c(-3,3),colors=c("blue","white","red"),breaks=c(-2,0,2),oob = scales::squish)+
    labs(title=names(sample_types)[h],fill="logFC")+
    guides(alpha="none",color="none")+
    theme_bw()+
    theme(axis.title.y = element_blank(),
          strip.background.y = element_blank(),
          strip.text.y.right = element_text(hjust=0,angle = 0),
          legend.position = "bottom")

  plot(ko.plots[[2]])  
}
```


# AMR DE analysis

## Plot mupirocin specific heatmap
```{r,fig.width = 11,fig.height=3}
library(ComplexHeatmap)

load("X:/mchung/leopoldo_segal_07/counts_conqur_KO.R")

counts.raw <- counts.ko$MEGARes
counts.raw <- counts.raw[!(grepl("RequiresSNPConfirmation",rownames(counts.raw))),]

counts.amr <- counts.raw
cpm.amr <- counts.amr/colSums(counts.amr) * 10^6
cpm.mupi <- cpm.amr[grep("Mupirocin",rownames(cpm.amr)),]
rownames(cpm.mupi) <- gsub("[|].*[|]",": ",rownames(cpm.mupi))

mat <- as.matrix(log2(cpm.mupi+1))


column_ha = HeatmapAnnotation("Group" = metadata$is_immunosuppressed,
                              col = list(Group = c("N"="#377eb8","Y"="#e4211c"))) 
                               
Heatmap(mat,
        name="log2CPM",
        colorRamp2(c(0,15,30), c("#9bcec8", "#efe920", "#ee2623")),
        top_annotation = column_ha,
        column_split = metadata$is_immunosuppressed,
        show_column_names = F)

pdf(paste0(WORKING.DIR,"/plots/Fig4d.pdf"),
    height=3,
    width=11)
Heatmap(mat,
        name="log2CPM",
        colorRamp2(c(0,15,30), c("#9bcec8", "#efe920", "#ee2623")),
        top_annotation = column_ha,
        column_split = metadata$is_immunosuppressed,
        show_column_names = F)
dev.off()
```
  
## Load counts
```{r}
load("X:/mchung/leopoldo_segal_07/counts_conqur_KO.R")

counts.raw <- counts.ko$MEGARes
counts.raw <- counts.raw[!(grepl("RequiresSNPConfirmation",rownames(counts.raw))),]

drug_categories <- strsplit(rownames(counts.raw),split="[|]")
drug_categories <- unlist(lapply(drug_categories,function(x){return(x[3])}))

counts.amr <- as.data.frame(matrix(nrow=length(unique(drug_categories)),
                                   ncol=ncol(counts.raw)))
rownames(counts.amr) <- unique(drug_categories)
colnames(counts.amr) <- colnames(counts.raw)
for(i in 1:nrow(counts.amr)){
  counts.amr[i,] <- colSums(counts.raw[drug_categories == rownames(counts.amr)[i],,drop=F])
}

counts.amr <- counts.amr[rowSums(counts.amr) >= 50,]
```

## Run edgeR

### Comparing Healthy v. Immunosuppressed

```{r}
edgeR.output <- list()
for(i in 1:length(samples)){
  sample_type <- gsub(" [|].*","",names(samples)[i])

  counts.subset <- counts.amr[,samples[[i]]]
  counts.subset <- counts.subset[,colSums(counts.subset) >= 100]
  metadata.subset <- metadata[match(colnames(counts.subset),metadata$Sample.ID),]
  
  imsupp <- factor(metadata.subset$is_immunosuppressed, levels=c("N","Y"))
  #weekpostintub <- ifelse(metadata.subset$weekpostintub >= 4,3,metadata.subset$weekpostintub)
  # run <- metadata.subset$run
  
  design <- model.matrix(~imsupp)
  # design <- model.matrix(~imsupp + run)
  
  y <- DGEList(counts=counts.subset,
               group=imsupp)
  keep <- filterByExpr(y)  
  y <- y[keep, , keep.lib.sizes = F]
  y <- calcNormFactors(y, method="TMM")
  y <- estimateDisp(y,design, robust=TRUE)
  fit <- glmQLFit(y,design,robust=TRUE)
    
  # qlf <- glmQLFTest(fit,coef=2:length(unique(groups)))
  qlf <- glmQLFTest(fit,coef=2)
  de <- topTags(qlf,n=Inf)$table
  edgeR.output[[i]] <- de
  
  print(length(rownames(edgeR.output[[i]])[edgeR.output[[i]]$FDR < 0.05 & !is.na(edgeR.output[[i]]$FDR)]))
}
names(edgeR.output) <- names(samples)
edgeR.imsupp <- edgeR.output
```


### Comparing Alive v. Dead

```{r}
edgeR.output <- list()
for(i in 1:length(samples)){
  sample_type <- gsub(" [|].*","",names(samples)[i])

  counts.subset <- counts.amr[,samples[[i]]]
  counts.subset <- counts.subset[,colSums(counts.subset) >= 100]
  metadata.subset <- metadata[match(colnames(counts.subset),metadata$Sample.ID),]
  
  doa <- factor(metadata.subset$dead_or_alive, levels=c("Alive","Dead"))
  #weekpostintub <- ifelse(metadata.subset$weekpostintub >= 4,3,metadata.subset$weekpostintub)
  # run <- metadata.subset$run
  
  design <- model.matrix(~doa)
  # design <- model.matrix(~imsupp + run)
  
  y <- DGEList(counts=counts.subset,
               group=doa)
  keep <- filterByExpr(y)  
  y <- y[keep, , keep.lib.sizes = F]
  y <- calcNormFactors(y, method="TMM")
  y <- estimateDisp(y,design, robust=TRUE)
  fit <- glmQLFit(y,design,robust=TRUE)
    
  # qlf <- glmQLFTest(fit,coef=2:length(unique(groups)))
  qlf <- glmQLFTest(fit,coef=2)
  de <- topTags(qlf,n=Inf)$table
  edgeR.output[[i]] <- de
  
  print(length(rownames(edgeR.output[[i]])[edgeR.output[[i]]$FDR < 0.05 & !is.na(edgeR.output[[i]]$FDR)]))
}
names(edgeR.output) <- names(samples)
edgeR.outcome <- edgeR.output
```


## Plot comparison dotplots

### Compare DE results between Healthy v. Immunosuppressed results as dotplot
```{r,fig.height=5,fig.width=10}
edgeR.output <- edgeR.imsupp

amr.plots <- list()
for(h in 1:length(sample_types)){
  sample_sets <- grep(names(sample_types)[h],names(samples))
  
  taxa <- c()
  logFC <- c()
  for(i in sample_sets){
    taxa <- c(taxa,rownames(edgeR.output[[i]]))
    logFC <- c(logFC,edgeR.output[[i]]$logFC)
  }
  score <- ifelse(logFC > 0,1,-1)
  taxa_stats <- as.data.frame(cbind(aggregate(score, by=list(Category=taxa), FUN=sum),
                                    aggregate(logFC, by=list(Category=taxa), FUN=mean)))
  taxa_stats <- taxa_stats[,c(1,2,4)]
  top_taxa <- taxa_stats[order(-abs(taxa_stats[,2]),-abs(taxa_stats[,3])),1] 
  
  taxa <- rev(top_taxa[top_taxa %in% taxa][1:50])
  taxa <- taxa[!is.na(taxa)]
  
  logFC.df <- as.data.frame(matrix(nrow=length(taxa),
                                   ncol=length(sample_sets)))
  rownames(logFC.df) <- taxa
  colnames(logFC.df) <- paste0("Week ",1:length(sample_sets))
  logCPM.df <- logFC.df
  sig.df <- logFC.df
  for(i in 1:length(sample_sets)){
    logFC.df[,i] <- edgeR.output[[sample_sets[i]]]$logFC[match(rownames(logFC.df),rownames(edgeR.output[[sample_sets[i]]]))]
    logCPM.df[,i] <- edgeR.output[[sample_sets[i]]]$logCPM[match(rownames(logCPM.df),rownames(edgeR.output[[sample_sets[i]]]))]
    sig.df[,i] <- ifelse(rownames(logFC.df) %in% rownames(edgeR.output[[sample_sets[i]]])[edgeR.output[[sample_sets[i]]]$FDR < 0.05 & !is.na(edgeR.output[[sample_sets[i]]]$FDR)],T,F)
  }
  
  plot.df <- reshape2::melt(as.matrix(logFC.df))
  colnames(plot.df) <- c("taxa","week","logFC")
  # plot.df$taxa <- factor(plot.df$taxa,levels=rownames(logCPM.df)[order(rowMeans(logCPM.df))])
  plot.df$logCPM <- reshape2::melt(as.matrix(logCPM.df))[,3]
  plot.df$sig <- ifelse(reshape2::melt(as.matrix(sig.df))[,3] == T, plot.df$logFC,NA)
  plot.df$kingdom <- annotation$superkingdom[match(plot.df$taxa,annotation$taxa)]
  plot.df$color <- ifelse(is.na(plot.df$sig),NA,"black")
  plot.df$alpha <- as.numeric(as.character(ifelse(is.na(plot.df$sig),0.2,1)))
  
  plot.df$logCPM <- as.numeric(ifelse(is.na(plot.df$logCPM),0,plot.df$logCPM))

  amr.plots[[h]] <- ggplot(plot.df,aes(x=logFC,y=taxa,fill=sig,color=color,alpha=alpha,size=logCPM))+
    geom_segment(aes(xend=0),alpha=0.4,color="black",size=0.7)+
    geom_point(pch=21,stat="identity")+
    geom_vline(aes(xintercept = 0),
               lty="dashed")+
    facet_grid(.~week, scales="free_y",space = "free")+
    scale_color_manual(values = c("black",NA))+
    scale_size_area(limits=c(0,20))+
    scale_fill_gradientn(limits = c(-3,3),colors=c("blue","white","red"),breaks=c(-2,0,2),oob = scales::squish)+
    labs(title=names(sample_types)[h],fill="logFC")+
    guides(alpha="none",color="none")+
    theme_bw()+
    theme(axis.title.y = element_blank(),
          strip.background.y = element_blank(),
          strip.text.y.right = element_text(hjust=0,angle = 0),
          legend.position = "bottom")

  plot(amr.plots[[h]])  
}
```

# Plot final figure
```{r,fig.height=8,fig.width=11}
egg::ggarrange(ko.plots[[1]]+
                 guides(fill="none",size="none")+
                 theme(plot.title = element_blank()),
               ko.plots[[2]]+
                 theme(legend.position = "right")+
                 theme(plot.title = element_blank()),
               amr.plots[[1]]+
                 guides(fill="none",size="none")+
                 theme(plot.title = element_blank()),
               heights=c(7.5,6.5,5.5),
               labels=LETTERS[1:3],
               ncol=1)

pdf(paste0(WORKING.DIR,"/plots/Fig4abc.pdf"),
    height=8,
    width=11)
egg::ggarrange(ko.plots[[1]]+
                 guides(fill="none",size="none")+
                 theme(plot.title = element_blank()),
               ko.plots[[2]]+
                 theme(legend.position = "right")+
                 theme(plot.title = element_blank()),
               amr.plots[[1]]+
                 guides(fill="none",size="none")+
                 theme(plot.title = element_blank()),
               heights=c(7.5,6.5,5.5),
               labels=LETTERS[1:3],
               ncol=1)
dev.off()
```
# Taxa DE analysis

## Run edgeR

### Comparing Healthy v. Immunosuppressed

```{r}
edgeR.output <- list()
for(i in 1:length(samples)){
  sample_type <- gsub(" [|].*","",names(samples)[i])
  taxa <- filter_taxa[[sample_type]]
    
  counts.subset <- counts[taxa,samples[[i]]]
  metadata.subset <- metadata[match(samples[[i]],metadata$Sample.ID),]
  
  imsupp <- factor(metadata.subset$is_immunosuppressed, levels=c("N","Y"))
  #weekpostintub <- ifelse(metadata.subset$weekpostintub >= 4,3,metadata.subset$weekpostintub)
  # run <- metadata.subset$run
  
  design <- model.matrix(~imsupp)
  # design <- model.matrix(~imsupp + run)
  
  y <- DGEList(counts=counts.subset,
               group=imsupp)
  keep <- filterByExpr(y)  
  y <- y[keep, , keep.lib.sizes = F]
  y <- calcNormFactors(y, method="TMM")
  y <- estimateDisp(y,design, robust=TRUE)
  fit <- glmQLFit(y,design,robust=TRUE)
    
  # qlf <- glmQLFTest(fit,coef=2:length(unique(groups)))
  qlf <- glmQLFTest(fit,coef=2)
  de <- topTags(qlf,n=Inf)$table
  edgeR.output[[i]] <- de
  
  print(length(rownames(edgeR.output[[i]])[edgeR.output[[i]]$FDR < 0.05 & !is.na(edgeR.output[[i]]$FDR)]))
}
names(edgeR.output) <- names(samples)
edgeR.imsupp <- edgeR.output
```



### Comparing Alive v. Dead

```{r}
edgeR.output <- list()
for(i in 1:length(samples)){
  sample_type <- gsub(" [|].*","",names(samples)[i])
  taxa <- filter_taxa[[sample_type]]
    
  counts.subset <- counts[taxa,samples[[i]]]
  counts.subset <- counts.subset[rowSums(counts.subset) != 0,]
  metadata.subset <- metadata[match(samples[[i]],metadata$Sample.ID),]
  
  doa <- factor(metadata.subset$dead_or_alive, levels=c("Alive","Dead"))
  #weekpostintub <- ifelse(metadata.subset$weekpostintub >= 4,3,metadata.subset$weekpostintub)
  # run <- metadata.subset$run
  
  design <- model.matrix(~doa)
  # design <- model.matrix(~imsupp + run)
  
  y <- DGEList(counts=counts.subset,
               group=doa)
  keep <- filterByExpr(y)  
  y <- y[keep, , keep.lib.sizes = F]
  y <- calcNormFactors(y, method="TMM")
  y <- estimateDisp(y,design, robust=TRUE)
  fit <- glmQLFit(y,design,robust=TRUE)
    
  # qlf <- glmQLFTest(fit,coef=2:length(unique(groups)))
  qlf <- glmQLFTest(fit,coef=2)
  de <- topTags(qlf,n=Inf)$table
  edgeR.output[[i]] <- de
  
  print(length(rownames(edgeR.output[[i]])[edgeR.output[[i]]$FDR < 0.05 & !is.na(edgeR.output[[i]]$FDR)]))
}
names(edgeR.output) <- names(samples)
edgeR.outcome <- edgeR.output
```


## Plot UpsetR of DE taxa
```{r}
up.imsupp <- list()
down.imsupp <- list()
up.outcome <- list()
down.outcome <- list()
for(i in 1:length(edgeR.imsupp)){
  up.imsupp[[i]] <- rownames(edgeR.imsupp[[i]])[edgeR.imsupp[[i]]$FDR < 0.05 & edgeR.imsupp[[i]]$logFC > 0]
  down.imsupp[[i]] <- rownames(edgeR.imsupp[[i]])[edgeR.imsupp[[i]]$FDR < 0.05 & edgeR.imsupp[[i]]$logFC < 0]
  
  up.outcome[[i]] <- rownames(edgeR.outcome[[i]])[edgeR.outcome[[i]]$FDR < 0.05 & edgeR.outcome[[i]]$logFC > 0]
  down.outcome[[i]] <- rownames(edgeR.outcome[[i]])[edgeR.outcome[[i]]$FDR < 0.05 & edgeR.outcome[[i]]$logFC < 0]
}
names(up.imsupp) <- names(edgeR.imsupp)
names(down.imsupp) <- names(edgeR.imsupp)
names(up.outcome) <- names(edgeR.outcome)
names(down.outcome) <- names(edgeR.outcome)
```


```{r,fig.height=3,fig.width=5}
ymax <- 25
scalemax <- 40

pdf(paste0(WORKING.DIR,"/plots/Fig3ab.pdf"),
    height=3,
    width=5)
m <- fromList(up.imsupp[1:3])
upset(m,sets=rev(names(m)),order.by = "freq",keep.order=T,mainbar.y.max=ymax, set_size.scale_max = scalemax,set_size.show=T,empty.intersections=T)
m <- fromList(down.imsupp[1:3])
upset(m,sets=rev(names(m)),order.by = "freq",keep.order=T,mainbar.y.max=ymax, set_size.scale_max = scalemax,set_size.show=T,empty.intersections=T)

m <- fromList(up.outcome[1:3])
upset(m,sets=rev(names(m)),order.by = "freq",keep.order=T,mainbar.y.max=ymax, set_size.scale_max = scalemax,set_size.show=T,empty.intersections=T)
m <- fromList(down.outcome[1:3])
upset(m,sets=rev(names(m)),order.by = "freq",keep.order=T,mainbar.y.max=ymax, set_size.scale_max = scalemax,set_size.show=T,empty.intersections=T)
dev.off()
```


## Plot comparison dotplots

### Compare DE results between Alive v. Dead results in Immuno as scatterplot
```{r,fig.height=6,fig.width=16}
xlab <- "logFC outcome"
edgeR.1 <- edgeR.outcome

ylab <- "logFC outcome in immunocompetent"
edgeR.2 <- edgeR.imsupp

up.labs <- sort(-table(c(unlist(up.imsupp[1:3]),
                         unlist(up.outcome[1:3]))))
up.labs <- names(up.labs)[abs(up.labs) >= 2]

down.labs <- sort(-table(c(unlist(down.imsupp[1:3]),
                         unlist(down.outcome[1:3]))))
down.labs <- names(down.labs)[abs(down.labs) >= 2]

labs <- c(up.labs,down.labs)

plot.list <- c()
for(i in 1:length(samples)){
  taxa <- unique(c(rownames(edgeR.1[[i]]),rownames(edgeR.2[[i]])))
  
  x <- edgeR.1[[i]]$logFC[match(taxa,rownames(edgeR.1[[i]]))]
  sig.x <- edgeR.1[[i]]$FDR[match(taxa,rownames(edgeR.1[[i]]))]
  
  y <- edgeR.2[[i]]$logFC[match(taxa,rownames(edgeR.2[[i]]))]
  sig.y <- edgeR.2[[i]]$FDR[match(taxa,rownames(edgeR.2[[i]]))]

  counts.subset <- DGEList(counts = counts[,samples[[i]]])
  cpm <- aveLogCPM(counts.subset)
  
  plot.df <- as.data.frame(cbind(taxa,x,sig.x,y,sig.y))
  plot.df$cpm <- cpm[match(taxa,rownames(counts))]
  
  plot.df$x <- as.numeric(as.character(plot.df$x))
  plot.df$sig.x <- as.numeric(as.character(plot.df$sig.x))

  plot.df$y <- as.numeric(as.character(plot.df$y))
  plot.df$sig.y <- as.numeric(as.character(plot.df$sig.y))

  plot.df <- plot.df[order(abs(rowMeans(plot.df[,c(2,4)])),decreasing=T),]
  plot.df$taxa <- factor(plot.df$taxa,levels=plot.df$taxa)

  plot.df$label <- ifelse(plot.df$sig.x < 0.05 & plot.df$sig.y < 0.05,
                          as.character(plot.df$taxa),NA)
  labs <- plot.df$label[!is.na(plot.df$label)][1:15]
  
  rm(taxa)
  plot.list[[i]] <- ggplot(plot.df,aes(x=x,y=y))+
    geom_vline(aes(xintercept=0),lty="dashed")+
    geom_hline(aes(yintercept=0),lty="dashed")+
    geom_point(alpha=ifelse(plot.df$sig.x < 0.05 & plot.df$sig.y < 0.05,
                            0.5,
                            0),
               color=ifelse(plot.df$sig.x < 0.05 & plot.df$sig.y < 0.05,
                            ifelse(plot.df$x > 0 & plot.df$y > 0,
                                   "red",
                                   ifelse(plot.df$x < 0 & plot.df$y < 0,
                                         "blue",
                                         "grey")),
                           "black"),
               size=plot.df$cpm)+
    # geom_text_repel(label=ifelse(plot.df$sig.x < 0.05 & plot.df$sig.y < 0.05,
    #                              taxa,
    #                              NA),
    #                 size=4)+
    geom_text_repel(label=ifelse(plot.df$label %in% labs,plot.df$label,NA),
                    max.overlaps=50,
                    size=4)+
    # scale_size_continuous(breaks=seq(2,20,2))+
    coord_cartesian(xlim=c(-10,10),ylim=c(-10,10))+
    labs(title=names(samples)[i],
           x=xlab,
           y=ylab)+
    theme_bw()
}
egg::ggarrange(plot.list[[1]],plot.list[[2]],plot.list[[3]],
               nrow=1,
               labels=LETTERS[1:3])
egg::ggarrange(plot.list[[4]],plot.list[[5]],plot.list[[6]],
               nrow=1,
               labels=LETTERS[1:3])


pdf(paste0(WORKING.DIR,"/plots/Fig3c.pdf"),
    height=6,
    width=16)
egg::ggarrange(plot.list[[1]],plot.list[[2]],plot.list[[3]],
               nrow=1,
               labels=LETTERS[1:3])
dev.off()
```

### Plot dot/bar plots of top DE results by week

```{r,fig.height=15,fig.width=10}
# edgeR.output <- edgeR.imsupp
edgeR.output <- edgeR.outcome
# edgeR.output <- edgeR.outcome.immunocompetent
# edgeR.output <- edgeR.outcome.immunocompromised

for(h in 1:length(sample_types)){
  sample_sets <- grep(names(sample_types)[h],names(samples))
  
  taxa <- c()
  logFC <- c()
  for(i in sample_sets){
    taxa <- c(taxa,rownames(edgeR.output[[i]])[edgeR.output[[i]]$FDR < 0.05])
    logFC <- c(logFC,edgeR.output[[i]]$logFC[edgeR.output[[i]]$FDR < 0.05])
  }
  score <- ifelse(logFC > 0,1,-1)
  taxa_stats <- as.data.frame(cbind(aggregate(score, by=list(Category=taxa), FUN=sum),
                                    aggregate(logFC, by=list(Category=taxa), FUN=mean)))
  taxa_stats <- taxa_stats[,c(1,2,4)]
  top_taxa <- taxa_stats[order(-abs(taxa_stats[,2]),-abs(taxa_stats[,3])),1] 
  
  taxa <- list()
  for(i in sample_sets){
    taxa[[i]] <- rownames(edgeR.output[[i]])[edgeR.output[[i]]$FDR < 0.05 & !is.na(edgeR.output[[i]]$FDR)]
  }
  taxa <- unique(unlist(taxa))
  # taxa <- rev(top_taxa[top_taxa %in% taxa][1:50])
  # taxa <- rev(top_taxa[top_taxa %in% taxa])

  logFC.df <- as.data.frame(matrix(nrow=length(taxa),
                                   ncol=length(sample_sets)))
  rownames(logFC.df) <- taxa
  colnames(logFC.df) <- paste0("Week ",1:length(sample_sets))
  logCPM.df <- logFC.df
  sig.df <- logFC.df
  for(i in 1:length(sample_sets)){
    logFC.df[,i] <- edgeR.output[[sample_sets[i]]]$logFC[match(rownames(logFC.df),rownames(edgeR.output[[sample_sets[i]]]))]
    logCPM.df[,i] <- edgeR.output[[sample_sets[i]]]$logCPM[match(rownames(logCPM.df),rownames(edgeR.output[[sample_sets[i]]]))]
    sig.df[,i] <- ifelse(rownames(logFC.df) %in% rownames(edgeR.output[[sample_sets[i]]])[edgeR.output[[sample_sets[i]]]$FDR < 0.05 & !is.na(edgeR.output[[sample_sets[i]]]$FDR)],T,F)
  }
  
  plot.df <- reshape2::melt(as.matrix(logFC.df))
  colnames(plot.df) <- c("taxa","week","logFC")
  # plot.df$taxa <- factor(plot.df$taxa,levels=rownames(logCPM.df)[order(rowMeans(logCPM.df))])
  plot.df$logCPM <- reshape2::melt(as.matrix(logCPM.df))[,3]
  plot.df$sig <- ifelse(reshape2::melt(as.matrix(sig.df))[,3] == T, plot.df$logFC,NA)
  plot.df$kingdom <- annotation$superkingdom[match(plot.df$taxa,annotation$taxa)]
  plot.df$color <- ifelse(is.na(plot.df$sig),NA,"black")
  plot.df$alpha <- as.numeric(as.character(ifelse(is.na(plot.df$sig),0.2,1)))
  
  p1 <- ggplot(plot.df,aes(x=logFC,y=taxa,fill=sig,color=color,alpha=alpha,size=logCPM))+
    geom_segment(aes(xend=0),alpha=0.4,color="black",size=0.7)+
    geom_point(pch=21,stat="identity")+
    geom_vline(aes(xintercept = 0),
               lty="dashed")+
    facet_grid(kingdom~week, scales="free_y",space = "free")+
    scale_color_manual(values = c("black",NA))+
    scale_fill_gradient2(low="blue",high="red")+
    labs(title=names(sample_types)[h],fill="logFC")+
    guides(alpha="none",color="none")+
    theme_bw()+
    theme(axis.title.y = element_blank(),
          strip.background.y = element_blank(),
          strip.text.y.right = element_text(hjust=0,angle = 0),
          legend.position = "bottom")

  plot(p1)
  
  # pdf(paste0(WORKING.DIR,"/plots/SupFig5.pdf"),
  #     height=15,
  #     width=10)
  
  pdf(paste0(WORKING.DIR,"/plots/SupFig6.pdf"),
      height=15,
      width=10)
  plot(p1)
  dev.off()
}
```

```{r,fig.height=11,fig.width=16}
plot.list <- c()
for(i in 1:length(samples)){
  taxa <- unique(c(rownames(edgeR.outcome[[i]]),rownames(edgeR.imsupp[[i]])))
  x <- edgeR.outcome[[i]]$logFC[match(taxa,rownames(edgeR.outcome[[i]]))]
  sig.x <- edgeR.outcome[[i]]$FDR[match(taxa,rownames(edgeR.outcome[[i]]))]
  
  y <- edgeR.imsupp[[i]]$logFC[match(taxa,rownames(edgeR.imsupp[[i]]))]
  sig.y <- edgeR.imsupp[[i]]$FDR[match(taxa,rownames(edgeR.imsupp[[i]]))]

  counts.subset <- DGEList(counts = counts[,samples[[i]]])
  cpm <- aveLogCPM(counts.subset)
  
  plot.df <- as.data.frame(cbind(taxa,x,sig.x,y,sig.y))
  plot.df$cpm <- cpm[match(taxa,rownames(counts))]
  
  plot.df$x <- as.numeric(as.character(plot.df$x))
  plot.df$sig.x <- as.numeric(as.character(plot.df$sig.x))

  plot.df$y <- as.numeric(as.character(plot.df$y))
  plot.df$sig.y <- as.numeric(as.character(plot.df$sig.y))

  plot.list[[i]] <- ggplot(plot.df,aes(x=x,y=y,label=taxa))+
    geom_vline(aes(xintercept=0),lty="dashed")+
    geom_hline(aes(yintercept=0),lty="dashed")+
    geom_point(alpha=ifelse(plot.df$sig.x < 0.05 & plot.df$sig.y < 0.05,
                            0.5,
                            0.05),
               color=ifelse(plot.df$sig.x < 0.05 & plot.df$sig.y < 0.05,
                            ifelse(plot.df$x > 0 & plot.df$y > 0,
                                   "red",
                                   ifelse(plot.df$x < 0 & plot.df$y < 0,
                                         "blue",
                                         "black")),
                           "black"),
               size=plot.df$cpm)+
    geom_text_repel(label=ifelse(plot.df$sig.x < 0.05 & plot.df$sig.y < 0.05,
                                 taxa,
                                 NA),
                    size=4)+
    # scale_size_continuous(breaks=seq(2,20,2))+
    # coord_cartesian(xlim=c(-15,15),ylim=c(-15,15))+
    labs(title=names(samples)[i],
           x="logFC outcome",
           y="logFC immune status")+
    theme_bw()
}
egg::ggarrange(plots=plot.list,
               nrow=2,
               labels=LETTERS[1:6])
```

