# Preanalysis

## Load packages
```{r}
library(decontam)
library(ggplot2)
library(ggpubr)
library(ggtext)
library(matrixStats)

WORKING.DIR="Z:/mchung/leopoldo_segal_07/"
# WORKING.DIR="X:/mchung/leopoldo_segal_07/"

```

## Load upstream object
```{r}
load(paste0(WORKING.DIR,"/upstream.R"))
# load(paste0(WORKING.DIR,"/upstream22.R"))

```


# Find contaminants

## Set functions

```{r}
wilcox_contam_test <- function(relab,neg){
  threshold <- 0.05
  
  p.value <- apply(relab,1,function(x){return(wilcox.test(as.numeric(as.character(x[!neg])),
                                                          as.numeric(as.character(x[neg])),
                                                          alternative="less",exact=FALSE)$p.value)})
  fdr <- unlist(lapply(p.value,function(x){return(p.adjust(x, method="BH", n = length(p.value)))}))
  
  contam <- as.data.frame(cbind(p.value,
                                fdr,
                                ifelse(p.value < 0.05 ,T, F)))
  rownames(contam) <- rownames(relab)
  colnames(contam) <- c("p.value","fdr","contaminant")
  return(contam)
}

identify_contam <- function(counts,relab,comparison.list){
  contam <- list()
  print(length(comparison.list))

  for(i in 1:length(comparison.list)){
    contam[[i]] <- list()
    
    counts.subset <- counts[,unlist(comparison.list[[i]])]
    relab.subset <- relab[,unlist(comparison.list[[i]])]
    libsizes <- colSums(counts.subset)
    bkg.samples <- c(rep(F,length(unlist(comparison.list[[i]][1]))),
                     rep(T,length(unlist(comparison.list[[i]][2]))))
    
    contam[[i]]$prev <- isContaminant(t(as.matrix(counts.subset)),
                                       method="prevalence",
                                       neg=bkg.samples,
                                       threshold=0.5)
    contam[[i]]$freq <- isContaminant(t(as.matrix(counts.subset)),
                                     method="frequency",
                                     neg=bkg.samples,
                                     conc=libsizes,
                                     threshold=0.5)
    contam[[i]]$combined <- isContaminant(t(as.matrix(counts.subset)),
                                         method="combined",
                                         neg=bkg.samples,
                                         conc=libsizes,
                                         threshold=0.5)
    contam[[i]]$wilcox <- wilcox_contam_test(relab.subset,
                                             neg=bkg.samples)
  }
  names(contam) <- names(comparison.list)
  return(contam)
}

plot_contam_plot <- function(contam,contam.prev,contam.freq,contam.combined,taxa){
  plot.df <- as.data.frame(rbind(cbind("Mann-Whitney U test",rownames(contam)[contam$contaminant == T]),
                                 cbind("decontam prevalence",rownames(contam.prev)[contam.prev$contaminant == T]),
                                 cbind("decontam frequency",rownames(contam.freq)[contam.freq$contaminant == T]),
                                 cbind("decontam combined",rownames(contam.combined)[contam.combined$contaminant == T])))
  plot.df <- plot.df[plot.df[,2] %in% taxa,]
  # 
  # for(i in 1:length(taxa)){
  #   plot.df[,2] <- gsub(taxa[i],paste0(taxa[i], " (",i,")"),plot.df[,2])
  #   taxa[i] <- paste0(taxa[i], " (",i,")")
  # }
  plot.df[,1] <- factor(plot.df[,1],levels=c("Mann-Whitney U test","decontam prevalence","decontam frequency","decontam combined"))
  plot.df[,2] <- factor(plot.df[,2],levels=rev(taxa))
  plot <- ggplot()+
    geom_point(mapping=aes(x=!!plot.df[,1],y=!!plot.df[,2],color=!!plot.df[,1]),size=2)+
    guides(color = "none")+
    theme_bw()+  
    scale_color_discrete(drop=FALSE) +
    scale_x_discrete(drop = FALSE)+
    scale_y_discrete(drop = FALSE)+
    theme(axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          axis.text.x = element_text(angle = 45, hjust = 1))
  
  return(plot)
}

plot_top_taxa_facet_boxplot <- function(relab.df,order.df,
                               contam_method,
                               contam,
                               annotation.taxa){
  taxa <- rownames(order.df)[order(-rowMedians(as.matrix(order.df)))][1:30]
  
  plot.df <- reshape2::melt(as.matrix(relab.df[taxa,]))
  plot.df[,4] <- annotation.taxa$superkingdom[match(plot.df[,1],annotation.taxa$taxa)]
  plot.df[,5] <- paste0(plot.df[,1], " (",match(plot.df[,1],taxa),")")
  plot.df[,5] <- ifelse(plot.df[,1] %in% rownames(contam[[contam_method]])[contam[[contam_method]]$contaminant == 1],
                        paste0("<span style='color:red'>", plot.df[,5], "</span>"),
                        paste0("<span style='color:black'>", plot.df[,5], "</span>"))
  plot.df[,5] <- factor(plot.df[,5],levels=rev(plot.df[1:30,5]))

  plot <- ggplot(plot.df,aes(x=!!plot.df[,3],
                             y=!!plot.df[,5],
                             fill=!!plot.df[,4]))+
    geom_boxplot()+
    labs(x="rel. ab.",y="")+
    guides(fill="none")+
    facet_grid(plot.df[,4] ~ .,scales="free",space = "free")+
    theme_bw()+
    theme(axis.text.y=element_markdown())
  
  return(plot)
}

plot_top_taxa_boxplot <- function(relab.df,
                                  order.df,
                                  contam_method,
                                  contam,
                                  annotation.taxa){
  taxa <- rownames(order.df)[order(-rowMedians(as.matrix(order.df)))][1:50]
  
  rank <- rank(-rowMedians(as.matrix(relab.df)))
  names(rank) <- rownames(relab.df)
  
  plot.df <- reshape2::melt(as.matrix(relab.df[taxa,]))
  plot.df[,4] <- annotation.taxa$superkingdom[match(plot.df[,1],annotation.taxa$taxa)]
  plot.df[,5] <- paste0(plot.df[,1], " (",rank[taxa],")")
  plot.df[,5] <- ifelse(plot.df[,1] %in% rownames(contam[[contam_method]])[contam[[contam_method]]$contaminant == 1],
                        paste0("<span style='color:red'>", plot.df[,5], "</span>"),
                        paste0("<span style='color:black'>", plot.df[,5], "</span>"))
  plot.df[,5] <- factor(plot.df[,5],levels=rev(plot.df[1:50,5]))

  plot <- ggplot(plot.df,aes(x=!!plot.df[,3],
                             y=!!plot.df[,5]))+
    geom_boxplot()+
    labs(x="rel. ab.",y="")+
    guides(fill="none")+
    theme_bw()+
    theme(axis.text.y=element_markdown())

  return(plot)
}
```

## Calculate relative abundance

```{r}
counts <- counts$Bracken
annotation <- annotation$Bracken

relab <- as.data.frame(apply(counts,2,function(x){x/sum(x)*100}))
```

## Run decontam and Mann-Whitney decontam functions

### Run decontam
```{r,fig.height=13,fig.width=13}
contam <- list()
comparison.list <- list()

comparison.list$All_vs_BKG <- list(c(which(metadata$sample_type == "BAL"),
                                     which(metadata$sample_type == "TA"),
                                     which(metadata$sample_type == "UA")),
                                   which(metadata$sample_type == "Background"))
comparison.list$LA_vs_BKG <- list(c(which(metadata$sample_type == "BAL"),
                                    which(metadata$sample_type == "TA")),
                                  which(metadata$sample_type == "Background"))
comparison.list$UA_vs_BKG <- list(which(metadata$sample_type == "UA"),
                                  which(metadata$sample_type == "Background"))

contam <- identify_contam(counts,
                          relab,
                          comparison.list)
```

#### Add decontam lists for (LA or UA vs BKG) and (LA and UA vs BKG)
```{r,fig.height=13,fig.width=13}
contam$LA_or_UA_vs_BKG <- list()
contam$LA_and_UA_vs_BKG <- list()

for(i in 1:length(contam$All_vs_BKG)){
  contam$LA_or_UA_vs_BKG[[i]] <- as.data.frame(matrix(nrow=nrow(contam$All_vs_BKG[[i]]),
                                                      ncol=1))
  contam$LA_or_UA_vs_BKG[[i]][,1] <- ifelse(contam$LA_vs_BKG[[i]]$contaminant == T | contam$UA_vs_BKG[[i]]$contaminant == T,
                                            T,F)
  rownames(contam$LA_or_UA_vs_BKG[[i]]) <- rownames(contam$All_vs_BKG[[i]])
  colnames(contam$LA_or_UA_vs_BKG[[i]]) <- "contaminant"
  
  contam$LA_and_UA_vs_BKG[[i]] <- as.data.frame(matrix(nrow=nrow(contam$All_vs_BKG[[i]]),
                                                       ncol=1))
  contam$LA_and_UA_vs_BKG[[i]][,1] <- ifelse(contam$LA_vs_BKG[[i]]$contaminant == T & contam$UA_vs_BKG[[i]]$contaminant == T,
                                             T,F)
  rownames(contam$LA_and_UA_vs_BKG[[i]]) <- rownames(contam$All_vs_BKG[[i]])
  colnames(contam$LA_and_UA_vs_BKG[[i]]) <- "contaminant"
}

names(contam$LA_or_UA_vs_BKG) <- names(contam$All_vs_BKG)
names(contam$LA_and_UA_vs_BKG) <- names(contam$All_vs_BKG)
```

#### Plot decontam summary for representative taxa
```{r,fig.height=10,fig.width=10}
expected_taxa <- list(good=c("Staphylococcus aureus","Pseudomonas aeruginosa","Salmonella enterica",
                             "Metamycoplasma salivarium","Pseudomonas aeruginosa",
                             "Candida albicans","Rothia mucilaginosa","Klebsiella pneumoniae",
                             "Streptococcus constellatus","Streptococcus pneumoniae","Streptococcus agalactiae",
                             "Streptococcus mitis","Streptococcus anginosus","Streptococcus pyogenes",
                             "Veillonella atypica","Veillonella dispar","Veillonella nakazawae",
                             "Veillonella parvula","Veillonella sp. S12025-13",
                             "Prevotella jejuni","Prevotella melaninogenica","Prevotella oris",
                             "Prevotella sp. oral taxon 299","Prevotella denticola",
                             "SARS-CoV2"),
                      contam=c("Bacillus anthracis",
                               "Aspergillus oryzae","Yersinia pestis","Burkholderia pseudomallei",
                               "Flavobacterium sediminis","Flavobacterium columnare","Flavobacterium davisii",
                               "Flavobacterium kingsejongi","Flavobacterium anhuiense",
                               "Burkholderia thailandensis"))

expected_taxa <- lapply(expected_taxa,function(x){rev(sort(x))})

plot.df <- as.data.frame(matrix(nrow=0,
                                ncol=length(unlist(expected_taxa))))
for(i in 1:length(contam)){
  for(j in 1:length(contam[[i]])){
    plot.df <- as.data.frame(rbind(plot.df,
                                   contam[[i]][[j]]$contaminant[match(as.character(unlist(expected_taxa)),rownames(contam[[i]][[j]]))]))
    rownames(plot.df)[nrow(plot.df)] <- paste0(names(contam)[i]," | ",names(contam[[i]])[j])
  }
}
colnames(plot.df) <- as.character(unlist(expected_taxa))

plot.df <- reshape2::melt(as.matrix(plot.df))
plot.df[,4] <- ifelse(plot.df[,2] %in% expected_taxa$contam,
                      "known contaminant","expected taxa")

plot.df[,3][is.na(plot.df[,3])] <- 0
for(i in 1:nrow(plot.df)){
  if(plot.df[i,4] == "known contaminant"){
    if(plot.df[i,3] == 1){
      plot.df[i,5] <- "Correctly labeled contaminant"
    }else{
      plot.df[i,5] <- "Incorrectly labeled"
    }
  }else{
    if(plot.df[i,3] == 1){
      plot.df[i,5] <- "Incorrectly labeled"
    }else{
      plot.df[i,5] <- "Correctly labeled expected taxa"
    }
  }
}

#plot.df <- plot.df[plot.df[,5] != "Incorrectly labeled",]

plot1 <- ggplot(data=plot.df,aes(x=plot.df[,1],fill=plot.df[,5]))+
  geom_bar()+
  scale_fill_manual(values=c("red","darkblue","darkgrey"))+
  labs(fill="Group")+
  theme_bw()+
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank())

plot2 <- ggplot(data=plot.df,aes(x=plot.df[,1],y=plot.df[,2],fill=plot.df[,5]))+
  facet_grid(plot.df[,4] ~ . ,scales="free",space="free")+
  geom_tile(colour = "grey50")+
  scale_fill_manual(values=c("red","darkblue","darkgrey"))+
  guides(fill="none")+
  theme_bw()+
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

egg::ggarrange(plot1,plot2,
               nrow=2,
               heights = c(1,5),
               labels = LETTERS[1:2])
```


#### Plot decontam results
```{r,fig.height=13,fig.width=13}
title <- "contam method: LA and UA vs BKG"
comp <- "LA_and_UA_vs_BKG"

order.df <- relab[,which(metadata$sample_type != "Background")]
taxa <- rownames(order.df)[order(-rowMedians(as.matrix(order.df)))][1:100]
plot1 <- plot_contam_plot(contam[[comp]]$wilcox,
                 contam[[comp]]$prev,
                 contam[[comp]]$freq,
                 contam[[comp]]$combined,
                 taxa)+
  labs(title="Top Non-BKG Taxa")

order.df <- relab[,which(metadata$sample_type == "Background")]
taxa <- rownames(order.df)[order(-rowMedians(as.matrix(order.df)))][1:100]
plot2 <- plot_contam_plot(contam[[comp]]$wilcox,
                          contam[[comp]]$prev,
                          contam[[comp]]$freq,
                          contam[[comp]]$combined,
                          taxa)+
  labs(title="Top BKG Taxa")

plot <- egg::ggarrange(plot1,plot2,
                       ncol=2,
                       labels = c("A","B"))
annotate_figure(plot,
                top = text_grob(title,color="red",size = 10,x=0,hjust = 0))
```

#### Plot top taxa in each sample type

```{r,fig.height=8,fig.width=15}
title <- "contam method: LA and UA vs BKG | red taxa labeled as contaminants using decontam prevalence\n"
comp <- "LA_and_UA_vs_BKG"
contam.method <- "prev"

plot1 <- plot_top_taxa_facet_boxplot(relab.df = relab[,which(metadata$sample_type == "BAL" | 
                                                              metadata$sample_type == "TA")],
                                     order.df = relab[,which(metadata$sample_type == "BAL" | 
                                                                       metadata$sample_type == "TA")],
                                     contam_method = contam.method,
                                     contam = contam[[comp]],
                                     annotation)+
  labs(title="Top BAL+TA Taxa")

plot2 <- plot_top_taxa_facet_boxplot(relab.df = relab[,which(metadata$sample_type == "UA")],
                                     order.df = relab[,which(metadata$sample_type == "UA")],
                                     contam_method = contam.method,
                                     contam = contam[[comp]],
                                     annotation)+
  labs(title="Top UA Taxa")

plot3 <- plot_top_taxa_facet_boxplot(relab.df = relab[,which(metadata$sample_type == "Background")],
                                     order.df = relab[,which(metadata$sample_type == "Background")],
                                     contam_method = contam.method,
                                     contam = contam[[comp]],
                                     annotation)+
  labs(title="Top BKG Taxa")

plot <- egg::ggarrange(plot1,plot2,plot3,
                       nrow=1,
                       labels = LETTERS[1:3])

annotate_figure(plot,
                top = text_grob(title,color="red",size = 10,x=0,hjust = 0))
```

#### Plot comparison box plots

```{r,fig.height=13,fig.width=15}
order.df <- relab[,which(metadata$sample_type == "BAL" |
                                   metadata$sample_type == "TA")]
  # order.df <- relab[,which(metadata$sample_type == "Background")]
title <- "contam method: LA and UA vs BKG | red taxa labeled as contaminants using decontam prevalence | ordered by top BAL+TA taxa\n"
comp <- "LA_and_UA_vs_BKG"
contam.method <- "prev"

plot1 <- plot_top_taxa_boxplot(relab.df = relab[,which(metadata$sample_type == "BAL" | 
                                                              metadata$sample_type == "TA")],
                               order.df = order.df,
                               contam_method = contam.method,
                               contam = contam[[comp]],
                               annotation)+
  geom_boxplot(fill="blue")+
  labs(title="BAL+TA Taxa")

plot2 <- plot_top_taxa_boxplot(relab.df = relab[,which(metadata$sample_type == "UA")],
                               order.df = order.df,
                               contam_method = contam.method,
                               contam = contam[[comp]],
                               annotation)+
  geom_boxplot(fill="orange")+
  labs(title="UA Taxa")


plot3 <- plot_top_taxa_boxplot(relab.df = relab[,which(metadata$sample_type == "Background")],
                               order.df = order.df,
                               contam_method = contam.method,
                               contam = contam[[comp]],
                               annotation)+
  geom_boxplot(fill="grey")+
  labs(title="BKG Taxa")


plot <- egg::ggarrange(plot1,plot2,plot3,
                       nrow=1,
                       labels = LETTERS[1:3])

pdf(paste0(WORKING.DIR,"/plots/contam.boxplot.pdf"),
    height=11,
    width=15,
    onefile = T)
annotate_figure(plot,
                top = text_grob(title,color="red",size = 10,x=0,hjust = 0))
dev.off()
```

#### Output contam summary table
```{r}
comp <- "LA_and_UA_vs_BKG"
contam.method <- "prev"

output.df <- contam[[comp]][[contam.method]]

output.df <- cbind(rownames(output.df),
                   output.df[,1])
colnames(output.df) <- c("taxa","is_contaminant")
output.df <- as.data.frame(output.df)

df <- relab[,which(metadata$sample_type == "BAL" | 
                             metadata$sample_type == "TA")]
output.df$LA_median_relab <- rowMedians(as.matrix(df))
output.df$LA_rank <- rank(-rowMedians(as.matrix(df)))

df <- relab[,which(metadata$sample_type == "UA")]
output.df$UA_median_relab <- rowMedians(as.matrix(df))
output.df$UA_rank <- rank(-rowMedians(as.matrix(df)))

df <- relab[,which(metadata$sample_type == "Background")]
output.df$BKG_median_relab <- rowMedians(as.matrix(df))
output.df$BKG_rank <- rank(-rowMedians(as.matrix(df)))

write.table(output.df,
            paste0(WORKING.DIR,"/tables/contam.",comp,".",contam.method,".tsv"),
            row.names = F,
            col.names = T,
            quote = F,
            sep = "\t")
```
