# Preanalysis

```{r}
library(edgeR)
library(igraph)
library(NetCoMi)
library(randomcoloR)
library(RColorBrewer)

WORKING.DIR="Z:/mchung/leopoldo_segal_07/"

load(paste0(WORKING.DIR,"/upstream.R"))
load(paste0(WORKING.DIR,"/de_analysis.R"))

counts <- counts$Bracken
annotation <- annotation$Bracken


```


```{r}
# counts <- read.delim(paste0(WORKING.DIR,"/tables/counts.conqur.bracken.tsv"))
counts <- apply(counts,2,function(x){as.numeric(as.character(x))})
rownames(counts) <- annotation$taxa

# metadata <- metadata[!is.na(metadata$imsupp),]

relab <- as.data.frame(apply(counts,2,function(x){x/sum(x)*100}))
```

## Set single sample list

```{r}
one_sample_per_subject <- function(metadata,sample_types, week){
  metadata <- metadata[metadata$sample_type %in% sample_types,]
  metadata <- metadata[metadata$weekpostintub %in% week,]
  metadata <- metadata[!duplicated(metadata$subject_id),]
  return(metadata$Sample.ID)
}

# samples <- list("BAL_and_TA | Week 1" = one_sample_per_subject(metadata, c("BAL","TA"), week = c(1)),
#                 "BAL_and_TA | Week 2" = one_sample_per_subject(metadata, c("BAL","TA"), week = c(2)),
#                 "BAL_and_TA | Week 3" = one_sample_per_subject(metadata, c("BAL","TA"), week = c(3)),
#                 "BAL_and_TA | Week 4" = one_sample_per_subject(metadata, c("BAL","TA"), week = c(4)),
#                 "UA | Week 1" = one_sample_per_subject(metadata, c("UA"), week = c(1)),
#                 "UA | Week 2" = one_sample_per_subject(metadata, c("UA"), week = c(2)),
#                 "UA | Week 3" = one_sample_per_subject(metadata, c("UA"), week = c(3)),
#                 "UA | Week 4" = one_sample_per_subject(metadata, c("UA"), week = c(4)))

samples <- list("BAL_and_TA | Week 1" = one_sample_per_subject(metadata, c("BAL","TA"), week = c(1)),
                "BAL_and_TA | Week 2" = one_sample_per_subject(metadata, c("BAL","TA"), week = c(2)),
                "BAL_and_TA | Week 3+" = one_sample_per_subject(metadata, c("BAL","TA"), week = c(3,4)),
                "UA | Week 1" = one_sample_per_subject(metadata, c("UA"), week = c(1)),
                "UA | Week 2" = one_sample_per_subject(metadata, c("UA"), week = c(2)),
                "UA | Week 3+" = one_sample_per_subject(metadata, c("UA"), week = c(3,4)))
```

### Get list of top taxa per sample type
```{r}
top_taxa <- list()
for(i in 1:length(samples)){
  sample_type <- gsub(" [|].*","",names(samples)[i])
  
  df <- counts[,samples[[i]]]
  cpm <- cpm(DGEList(df))
  logCPM <- cpm(cpm, log=TRUE, prior.count = 1)  
  taxa <- rownames(df)[order(-rowMeans(logCPM))]
  taxa <- taxa[taxa != "Equine infectious anemia virus"]
  
  kingdom <- annotation$superkingdom[match(taxa,annotation$taxa)]
  subkingdom <- annotation$kingdom[match(taxa,annotation$taxa)]

  top_taxa[[sample_type]] <- unique(c(top_taxa[[sample_type]],taxa[1:100]))
}
```

# Run NetCoMi

## Compare each week using top abundant taxa

### Construct network for single covariates
```{r}
net_spring <- list()
for(i in 1:length(samples)){
  sample_type <- gsub(" [|].*","",names(samples)[i])

  metadata.subset <- metadata[match(samples[[i]],metadata$Sample.ID),]
  
  # covariate <- "is_immunosuppressed"
  # groups <- c("N","Y")
  # df1 <- counts[,metadata.subset$Sample.ID[metadata.subset$is_immunosuppressed == groups[1]]]
  # df2 <- counts[,metadata.subset$Sample.ID[metadata.subset$is_immunosuppressed == groups[2]]]
  # 
  covariate <- "dead_or_alive"
  groups <- c("Alive","Dead")
  df1 <- counts[,metadata.subset$Sample.ID[metadata.subset$dead_or_alive == groups[1]]]
  df2 <- counts[,metadata.subset$Sample.ID[metadata.subset$dead_or_alive == groups[2]]]
  
  
  df1 <- t(df1[top_taxa[[sample_type]],])
  df2 <- t(df2[top_taxa[[sample_type]],])
  
  net_spring[[i]] <- netConstruct(data = df1, 
                                  data2 = df2,  
                                  # filtTaxPar = list(highestFreq = 50),
                                  # filtSamp = "totalReads",
                                  # filtSampPar = list(totalReads = 1000),
                                  measure = "spring",
                                  measurePar = list(nlambda=10, 
                                                    rep.num=10),
                                  normMethod = "clr", 
                                  zeroMethod = "pseudoZO",
                                  sparsMethod = "threshold", 
                                  thresh = 0.3,
                                  dissFunc = "signed",
                                  verbose = 2,
                                  seed = 123456)
  print(i)
}
names(net_spring) <- names(samples)

save(net_spring,
     file=paste0(WORKING.DIR,"/netcomi.",covariate,".R"))
```


### Construct network for healthy/immunosuppressed subsets
```{r}
net_spring <- list()
for(i in 4:length(samples)){
  sample_type <- gsub(" [|].*","",names(samples)[i])

  metadata.subset <- metadata[match(samples[[i]],metadata$Sample.ID),]
  
  imsupp <- "immunocompetent"
  covariate <- "dead_or_alive"
  groups <- c("Alive","Dead")
  df1 <- counts[,metadata.subset$Sample.ID[metadata.subset$is_immunosuppressed == "N" & metadata.subset$dead_or_alive == groups[1]]]
  df2 <- counts[,metadata.subset$Sample.ID[metadata.subset$is_immunosuppressed == "N" & metadata.subset$dead_or_alive == groups[2]]]
  
  # imsupp <- "immunocompromised"
  # covariate <- "dead_or_alive"
  # groups <- c("Alive","Dead")
  # df1 <- counts[,metadata.subset$Sample.ID[metadata.subset$is_immunosuppressed == "Y" & metadata.subset$dead_or_alive == groups[1]]]
  # df2 <- counts[,metadata.subset$Sample.ID[metadata.subset$is_immunosuppressed == "Y" & metadata.subset$dead_or_alive == groups[2]]]
  
  
  df1 <- t(df1[top_taxa[[sample_type]],])
  df2 <- t(df2[top_taxa[[sample_type]],])
  
  net_spring[[i]] <- netConstruct(data = df1, 
                                  data2 = df2,  
                                  # filtTaxPar = list(highestFreq = 50),
                                  # filtSamp = "totalReads",
                                  # filtSampPar = list(totalReads = 1000),
                                  measure = "spring",
                                  measurePar = list(nlambda=10, 
                                                    rep.num=10),
                                  normMethod = "clr", 
                                  zeroMethod = "pseudoZO",
                                  sparsMethod = "threshold", 
                                  thresh = 0.3,
                                  dissFunc = "signed",
                                  verbose = 2,
                                  seed = 123456)
  print(i)
}
names(net_spring) <- names(samples)

save(net_spring,
     file=paste0(WORKING.DIR,"/netcomi.",imsupp,".",covariate,".R"))
```

## Plot comparison network using NetCoMi functions
```{r, fig.height=20,fig.width=40}
plot(net_spring[[i]]$assoMat1)
p <- plot(props_net_spring, 
          sameLayout = TRUE, 
          layout="layout_with_kk",
          layoutGroup = 1,
          rmSingles = F, 
          nodeColor = "cluster", 
          nodeSize = "mclr",
          repulsion = 5,
          labelScale = FALSE,
          cexLabels = 2,
          nodeSizeSpread = 3,
          cexNodes = 2,
          hubBorderCol = "gray40",
          title1 = "Network on OTU level with SPRING associations", 
          groupNames = groups,
          showTitle = TRUE,
          cexTitle = 2.3)
legend("bottom", title = "estimated association:", legend = c("+","-"), 
       col = c("#009900","red"), inset = 0.02, cex = 4, lty = 1, lwd = 4, 
       bty = "n", horiz = TRUE)
```

## Plot comparison networks with igraph
```{r, fig.height=10,fig.width=20}
covariate <- "dead_or_alive"
covar.groups <- c("Alive","Dead")

# covariate <- "is_immunosuppressed"
# covar.groups <- c("Immunocompetent","Immunosuppressed")

load(paste0(WORKING.DIR,"/netcomi.",covariate,".R"))

select_nodes <- c("SARS-CoV2","Candida albicans","Proteus mirabilis",
                  grep("Staphylococcus",top_taxa$BAL_and_TA,value=T),
                  grep("Bacillus",top_taxa$BAL_and_TA,value=T),
                  grep("Corynebacterium",top_taxa$BAL_and_TA,value=T),
                  grep("Lactobacillus",top_taxa$BAL_and_TA,value=T),
                  grep("Streptococcus",top_taxa$BAL_and_TA,value=T),
                  grep("Prevotella",top_taxa$BAL_and_TA,value=T))
for(i in 1:length(net_spring)){
  diff_net_spring <- diffnet(net_spring[[i]],
                             diffMethod = "fisherTest", 
                             adjust = "none")

  props_net_spring <- netAnalyze(net_spring[[i]], 
                                 clustMethod = "cluster_fast_greedy",
                                 weightDeg = TRUE,
                                 normDeg = FALSE,
                                 gcmHeat = FALSE)
  
  graphA <- igraph::graph_from_adjacency_matrix(net_spring[[i]]$adjaMat1, weighted = TRUE)
  graphA <- simplify(graphA,remove.multiple = T, remove.loops = T)
  set.seed(123456)
  lay_fr <- igraph::layout_with_kk(graphA)
  rownames(lay_fr) <- rownames(net_spring[[i]]$adjaMat1)
  graphA <- as.undirected(graphA)
  
  graphB <- igraph::graph_from_adjacency_matrix(net_spring[[i]]$adjaMat2, weighted = TRUE)
  graphB <- simplify(graphB,remove.multiple = T, remove.loops = T)
  set.seed(123456)
  # lay_fr <- igraph::layout_with_kk(graphB)
  # rownames(lay_fr) <- rownames(net_spring[[i]]$adjaMat1)
  graphB <- as.undirected(graphB)
  
  clusters <- c(paste0("clust1_",sort(unique(props_net_spring$clustering$clust1))),
                paste0("clust2_",sort(unique(props_net_spring$clustering$clust2))))
  set.seed(123456)
  color <- distinctColorPalette(length(clusters))
  for(j in 1:length(clusters)){
    if(gsub("_.*","",clusters[j]) == "clust1"){
      props_net_spring$clustering$clust1 <- ifelse(props_net_spring$clustering$clust1 == gsub(".*_","",clusters[j]),
                                             color[j],
                                             props_net_spring$clustering$clust1)
    }else{
      props_net_spring$clustering$clust2 <- ifelse(props_net_spring$clustering$clust2 == gsub(".*_","",clusters[j]),
                                             color[j],
                                             props_net_spring$clustering$clust2)
    }
  }
  kingdom <- annotation$superkingdom[match(names(V(graphA)),annotation$taxa)] 

  ends1 <- ends(graphA,es=E(graphA))
  ends1$key <- ifelse(ends1[,2]>ends1[,1],paste0(ends1[,1],"--",ends1[,2]),paste0(ends1[,2],"--",ends1[,1]))
  net_spring[[i]]$edgelist1$key <- ifelse(net_spring[[i]]$edgelist1$v2 > net_spring[[i]]$edgelist1$v1,
                                          paste0(net_spring[[i]]$edgelist1$v1,"--",net_spring[[i]]$edgelist1$v2),
                                          paste0(net_spring[[i]]$edgelist1$v2,"--",net_spring[[i]]$edgelist1$v1))
  V(graphA)$color <- props_net_spring$clustering$clust1
  V(graphA)$shape <- ifelse(kingdom == "Bacteria","circle","csquare")
  V(graphA)$size <- log2(colMeans(net_spring[[i]]$normCounts1))
  V(graphA)$size <- ifelse(V(graphA)$size > 1,V(graphA)$size,1)*0.8
  E(graphA)$color <- ifelse(net_spring[[i]]$edgelist1$asso[match(ends1$key,net_spring[[i]]$edgelist1$key)] > 0,
                            "gray60","indianred")
  print(table(E(graphA)$color))
  # E(graphA)$color <- "darkgrey"
  E(graphA)$width <- abs(net_spring[[i]]$edgelist1$asso[match(ends1$key,net_spring[[i]]$edgelist1$key)])*20
  
  # V(graphA)$name <- ifelse(V(graphA)$name %in% props_net_spring$hubs$hubs1,V(graphA)$name,NA)
  V(graphA)$name <- ifelse(V(graphA)$name %in% select_nodes,V(graphA)$name,NA)
  V(graphA)$name <- ifelse(V(graphA)$name == "SARS-CoV2","SARS\nCoV2",substr(V(graphA)$name, start = 1, stop = 3))
  
  plot(graphA,
     layout=lay_fr,
     vertex.label.cex = V(graphA)$size*0.1,
     vertex.label.color = "black",
     vertex.label.family = "Helvetica",
     main = covar.groups[1])
  
  pdf(paste0(WORKING.DIR,"/plots/NetCoMi/",gsub(" ","_",gsub(" [|] ","_",names(net_spring)[i])),
             ".individual_week_comparison.",covariate,".selectlabels.pdf"),
      height=10,
      width=20)
  # png(paste0(WORKING.DIR,"/plots/NetCoMi/",gsub(" ","_",gsub(" [|] ","_",names(net_spring)[i])),
  #            ".individual_week_comparison.",covariate,".selectlabels.png"),
  #     height=20,
  #     width=40,
  #     res = 300,
  #     units = "in")
  # 
  print("test")
  par(mfrow = c(1, 2))
  plot(graphA,
     layout=lay_fr,
     vertex.label.cex = V(graphA)$size*0.1,
     vertex.label.color = "black",
     vertex.label.family = "Helvetica",
     main = covar.groups[1])
  
  ends2 <- ends(graphB,es=E(graphB))
  ends2$key <- ifelse(ends2[,2]>ends2[,1],paste0(ends2[,1],"--",ends2[,2]),paste0(ends2[,2],"--",ends2[,1]))
  net_spring[[i]]$edgelist2$key <- ifelse(net_spring[[i]]$edgelist2$v2 > net_spring[[i]]$edgelist2$v1,
                                          paste0(net_spring[[i]]$edgelist2$v1,"--",net_spring[[i]]$edgelist2$v2),
                                          paste0(net_spring[[i]]$edgelist2$v2,"--",net_spring[[i]]$edgelist2$v1))
  V(graphB)$color <- props_net_spring$clustering$clust2
  V(graphB)$shape <- ifelse(kingdom == "Bacteria","circle","csquare")
  V(graphB)$size <- log2(colMeans(net_spring[[i]]$normCounts2))
  V(graphB)$size <- ifelse(V(graphB)$size > 1,V(graphB)$size,1)*0.8
  E(graphB)$color <- ifelse(net_spring[[i]]$edgelist2$asso[match(ends2$key,net_spring[[i]]$edgelist2$key)] > 0,
                            "gray60","indianred")
  # E(graphB)$color <- "darkgrey"
  print(table(E(graphB)$color))
  E(graphB)$width <- abs(net_spring[[i]]$edgelist2$asso[match(ends2$key,net_spring[[i]]$edgelist2$key)])*20
  
  # V(graphB)$name <- ifelse(V(graphB)$name %in% props_net_spring$hubs$hubs1,V(graphB)$name,NA)
  V(graphB)$name <- ifelse(V(graphB)$name %in% select_nodes,V(graphB)$name,NA)
  V(graphB)$name <- ifelse(V(graphB)$name == "SARS-CoV2","SARS\nCoV2",substr(V(graphB)$name, start = 1, stop = 3))
  
  plot(graphB,
     layout=lay_fr,
     vertex.label.cex = V(graphB)$size*0.1,
     vertex.label.color = "black",
     vertex.label.family = "Helvetica",
     main = covar.groups[2])


  # for(j in 1:length(clusters)){
  #   if(gsub("_.*","",clusters[j]) == "clust1"){
  #     taxa <- names(props_net_spring$clustering$clust1)[props_net_spring$clustering$clust1 == color[j]]
  #   }else{
  #     taxa <- names(props_net_spring$clustering$clust2)[props_net_spring$clustering$clust2 == color[j]]
  #   }
  #   kingdom <- annotation$superkingdom[match(taxa,annotation$taxa)] 
  #     
  #   graphA.sub <- induced_subgraph(graphA, V(graphA)[taxa])
  #   graphB.sub <- induced_subgraph(graphB, V(graphB)[taxa])
  # 
  #   ends1 <- ends(graphA.sub,es=E(graphA.sub))
  #   ends1$key <- ifelse(ends1[,2]>ends1[,1],paste0(ends1[,1],"--",ends1[,2]),paste0(ends1[,2],"--",ends1[,1]))
  # 
  #   ends2 <- ends(graphB.sub,es=E(graphB.sub))
  #   ends2$key <- ifelse(ends2[,2]>ends2[,1],paste0(ends2[,1],"--",ends2[,2]),paste0(ends2[,2],"--",ends2[,1]))
  #   
  #   V(graphA.sub)$color <- props_net_spring$clustering$clust1[taxa]
  #   V(graphB.sub)$shape <- ifelse(kingdom == "Bacteria","circle","csquare")
  #   V(graphA.sub)$size <- log2(colMeans(net_spring[[i]]$normCounts1[,taxa,drop=F]))
  #   V(graphA.sub)$size <- ifelse(V(graphA.sub)$size > 1,V(graphA.sub)$size,1)
  #   E(graphA.sub)$color <- ifelse(net_spring[[i]]$edgelist1$asso[match(ends1$key,net_spring[[i]]$edgelist1$key)] > 0,
  #                                 "#006400","#ff3939")
  #   E(graphA.sub)$width <- abs(net_spring[[i]]$edgelist1$asso[match(ends1$key,net_spring[[i]]$edgelist1$key)])*100
  # 
  #   par(mfrow = c(1, 2))
  #   plot(graphA.sub,
  #        layout=lay_fr[taxa,,drop=F],
  #        vertex.label.cex = 2,
  #        vertex.label.color = "black",
  #        vertex.label.family = "Helvetica",
  #        rmSingles = F,
  #        main = groups[1])
  #   
  #   V(graphB.sub)$color <- props_net_spring$clustering$clust2[taxa]
  #   V(graphB.sub)$shape <- ifelse(kingdom == "Bacteria","circle","csquare")
  #   V(graphB.sub)$size <- log2(colMeans(net_spring[[i]]$normCounts2[,taxa,drop=F]))
  #   V(graphB.sub)$size <- ifelse(V(graphB.sub)$size > 1,V(graphB.sub)$size,1)
  #   E(graphB.sub)$color <- ifelse(net_spring[[i]]$edgelist2$asso[match(ends2$key,net_spring[[i]]$edgelist2$key)] > 0,
  #                                 "#006400","#ff3939")
  #   E(graphB.sub)$width <- abs(net_spring[[i]]$edgelist2$asso[match(ends2$key,net_spring[[i]]$edgelist2$key)])*100
  # 
  #   plot(graphB.sub,
  #        layout=lay_fr[taxa,,drop=F],
  #        vertex.label.cex = 2,
  #        vertex.label.color = "black",
  #        vertex.label.family = "Helvetica",
  #        rmSingles = F,
  #        main=groups[2])
  # }
  print("test2")
  dev.off()
}
```

```{r,fig.height=10,fig.width=20}
par(mfrow = c(1, 2))

plot(graphA,
     labelScale = T,
      posCol = "darkturquoise", 
     negCol = "orange",
     edgeTranspLow = 0,
     edgeTranspHigh = 40,
     layout=lay_fr,
     # vertex.label.cex = 2,
     # vertex.label.color = "black",
     # vertex.label.family = "Helvetica",
     rmSingles = F,
     main = covar.groups[1])

V(graphA)$color <- props_net_spring$clustering$clust1
V(graphA)$shape <- ifelse(kingdom == "Bacteria","circle","csquare")
V(graphA)$size <- log2(colMeans(net_spring[[i]]$normCounts1))
V(graphA)$size <- ifelse(V(graphA)$size > 1,V(graphA)$size,1)*0.8
E(graphA)$color <- ifelse(net_spring[[i]]$edgelist1$asso[match(ends1$key,net_spring[[i]]$edgelist1$key)] > 0,
                          "#0d4c00","#9d3c84")
E(graphA)$color <- "darkgrey"
E(graphA)$width <- abs(net_spring[[i]]$edgelist1$asso[match(ends1$key,net_spring[[i]]$edgelist1$key)])*20

# V(graphA)$name <- ifelse(V(graphA)$name %in% props_net_spring$hubs$hubs1,V(graphA)$name,NA)
V(graphA)$name <- ifelse(V(graphA)$name %in% select_nodes,V(graphA)$name,NA)
V(graphA)$name <- ifelse(V(graphA)$name == "SARS-CoV2","SARS-CoV2",substr(V(graphA)$name, start = 1, stop = 3))
  
plot(graphA,
     layout=lay_fr,
     vertex.label.cex = V(graphA)$size*0.1,
     vertex.label.color = "black",
     vertex.label.family = "Helvetica")
```


## Plot delta node degree and betweeness

### Calculate betweenness node degree
```{r,fig.height=6,fig.width=15}
plot.df <- as.data.frame(matrix(nrow=0,
                                ncol=3))

plot.list <- list()

for(i in 1:length(net_spring)){
  g1 <- igraph::graph_from_adjacency_matrix(net_spring[[i]]$adjaMat1, weighted = TRUE)
  g2 <- igraph::graph_from_adjacency_matrix(net_spring[[i]]$adjaMat2, weighted = TRUE)

  names <- V(g1)$name
  delta <- betweenness(g2) - betweenness(g1)
  
  plot.df <- as.data.frame(cbind(names,delta))
  
  plot.df$delta <- as.numeric(as.character(delta))
  
  top <- plot.df$names[order(-plot.df$delta)][1:15]
  bottom <- rev(plot.df$names[order(plot.df$delta)][1:15])
  
  plot.df <- plot.df[plot.df$names %in% c(top,bottom),]
  plot.df$names <- factor(plot.df$names,levels=rev(c(top,bottom)))
  
  plot.list[[i]] <- ggplot(plot.df,aes(x=delta,y=names,fill=delta))+
    geom_hline(yintercept = 15.5,lty="dashed",color="darkgrey")+
    geom_vline(xintercept = 0,lty="dashed")+
    geom_bar(stat="identity")+
    scale_fill_gradient2(low="blue",high="red")+
    labs(title=names(samples)[i],
         x="delta immunosuppressed-healthy node betweenness")+
    theme_bw()+
    theme(axis.title.y = element_blank())
}

egg::ggarrange(plot.list[[1]]+
                 theme(axis.title.x = element_blank()),
               plot.list[[2]],
               plot.list[[3]]+
                 theme(axis.title.x = element_blank()),
               nrow=1)
```

### Calculate delta node degree
```{r,fig.height=6,fig.width=15}
plot.df <- as.data.frame(matrix(nrow=0,
                                ncol=3))

plot.list <- list()

for(i in 1:length(net_spring)){
  g1 <- igraph::graph_from_adjacency_matrix(net_spring[[i]]$adjaMat1, weighted = TRUE)
  g2 <- igraph::graph_from_adjacency_matrix(net_spring[[i]]$adjaMat2, weighted = TRUE)

  names <- V(g1)$name
  delta <- degree(g2) - degree(g1)
  
  plot.df <- as.data.frame(cbind(names,delta))
  
  plot.df$delta <- as.numeric(as.character(delta))
  
  top <- plot.df$names[order(-plot.df$delta)][1:15]
  bottom <- rev(plot.df$names[order(plot.df$delta)][1:15])
  
  plot.df <- plot.df[plot.df$names %in% c(top,bottom),]
  plot.df$names <- factor(plot.df$names,levels=rev(c(top,bottom)))
  
  plot.list[[i]] <- ggplot(plot.df,aes(x=delta,y=names,fill=delta))+
    geom_hline(yintercept = 15.5,lty="dashed",color="darkgrey")+
    geom_vline(xintercept = 0,lty="dashed")+
    geom_bar(stat="identity")+
    scale_fill_gradient2(low="blue",high="red")+
    labs(title=names(samples)[i],
         x="delta immunosuppressed-healthy node degree")+
    theme_bw()+
    theme(axis.title.y = element_blank())
}

egg::ggarrange(plot.list[[1]]+
                 theme(axis.title.x = element_blank()),
               plot.list[[2]],
               plot.list[[3]]+
                 theme(axis.title.x = element_blank()),
               nrow=1)
```

## Conduct network attack based analyses

### Load NetCoMi object
```{r}
# covariate <- "dead_or_alive"
# covar.groups <- c("Alive","Dead")

covariate <- "is_immunosuppressed"
covar.groups <- c("Immunocompetent","Immunosuppressed")

load(paste0(WORKING.DIR,"/netcomi.",covariate,".R"))
```


### Betweenness-based attacks
```{r}
plot.df <- as.data.frame(matrix(nrow=0,
                                ncol=3))

for(i in 1:length(net_spring)){
  g <- igraph::graph_from_adjacency_matrix(net_spring[[i]]$adjaMat1, weighted = TRUE)
  components <- igraph::clusters(g, mode="weak")
  print(max(components$csize))
  
  nodes <- length(V(g)$name)
  comp_size <- max(components$csize)
    
  while(length(V(g)$name) > 5){
    set.seed(123)
    g <- delete.vertices(g, V(g)$name == names(sort(-betweenness(g)))[1])
    components <- igraph::clusters(g, mode="weak")
    
    nodes <- c(nodes,length(V(g)$name))
    comp_size <- c(comp_size,max(components$csize))
  }
  
  sample <- rep(paste0(names(net_spring)[i]," | ",covar.groups[1]),length(nodes))

  nodes <- (nodes[1]-nodes)/nodes[1]
  comp_size <- comp_size/comp_size[1]
  
  plot.df <- as.data.frame(rbind(plot.df,
                                 cbind(sample,
                                       nodes,
                                       comp_size)))
  #### 
  g <- igraph::graph_from_adjacency_matrix(net_spring[[i]]$adjaMat2, weighted = TRUE)
  components <- igraph::clusters(g, mode="weak")
  print(max(components$csize))
  
  nodes <- length(V(g)$name)
  comp_size <- max(components$csize)
    
  while(length(V(g)$name) > 5){
    set.seed(123)
    g <- delete.vertices(g, V(g)$name == names(sort(-betweenness(g)))[1])
    components <- igraph::clusters(g, mode="weak")
    
    nodes <- c(nodes,length(V(g)$name))
    comp_size <- c(comp_size,max(components$csize))
  }
  
  sample <- rep(paste0(names(net_spring)[i]," | ",covar.groups[2]),length(nodes))

  nodes <- (nodes[1]-nodes)/nodes[1]
  comp_size <- comp_size/comp_size[1]
  
  plot.df <- as.data.frame(rbind(plot.df,
                                 cbind(sample,
                                       nodes,
                                       comp_size)))
}

plot.df$nodes <- as.numeric(as.character(plot.df$nodes))
plot.df$comp_size <- as.numeric(as.character(plot.df$comp_size))

plot.df <- plot.df[!grepl("UA",plot.df$sample),]
# plot.df <- plot.df[grepl("UA",plot.df$sample),]

p1 <- ggplot(plot.df,aes(x=nodes,y=comp_size,color=sample))+
  geom_line(size=1)+
  labs(title="Betweenness-based attack",
       x="percentage of nodes removed",
       y="size of largest component",
       color="Network")+
  theme_bw()
```


### Degree-based attacks
```{r}
plot.df <- as.data.frame(matrix(nrow=0,
                                ncol=3))

for(i in 1:length(net_spring)){
  g <- igraph::graph_from_adjacency_matrix(net_spring[[i]]$adjaMat1, weighted = TRUE)
  components <- igraph::clusters(g, mode="weak")
  print(max(components$csize))
  
  nodes <- length(V(g)$name)
  comp_size <- max(components$csize)
    
  while(length(V(g)$name) > 5){
    set.seed(123)
    g <- delete.vertices(g, V(g)$name == names(sort(-degree(g)))[1])
    components <- igraph::clusters(g, mode="weak")
    
    nodes <- c(nodes,length(V(g)$name))
    comp_size <- c(comp_size,max(components$csize))
  }
  
  sample <- rep(paste0(names(net_spring)[i]," | ",covar.groups[1]),length(nodes))
  
  nodes <- (nodes[1]-nodes)/nodes[1]
  comp_size <- comp_size/comp_size[1]
  
  plot.df <- as.data.frame(rbind(plot.df,
                                 cbind(sample,
                                       nodes,
                                       comp_size)))
  #### 
  g <- igraph::graph_from_adjacency_matrix(net_spring[[i]]$adjaMat2, weighted = TRUE)
  components <- igraph::clusters(g, mode="weak")
  print(max(components$csize))
  
  nodes <- length(V(g)$name)
  comp_size <- max(components$csize)
    
  while(length(V(g)$name) > 5){
    set.seed(123)
    g <- delete.vertices(g, V(g)$name == names(sort(-degree(g)))[1])
    components <- igraph::clusters(g, mode="weak")
    
    nodes <- c(nodes,length(V(g)$name))
    comp_size <- c(comp_size,max(components$csize))
  }
  
  sample <- rep(paste0(names(net_spring)[i]," | ",covar.groups[2]),length(nodes))
    
  nodes <- (nodes[1]-nodes)/nodes[1]
  comp_size <- comp_size/comp_size[1]
  
  plot.df <- as.data.frame(rbind(plot.df,
                                 cbind(sample,
                                       nodes,
                                       comp_size)))
}

plot.df$nodes <- as.numeric(as.character(plot.df$nodes))
plot.df$comp_size <- as.numeric(as.character(plot.df$comp_size))

plot.df <- plot.df[!grepl("UA",plot.df$sample),]
# plot.df <- plot.df[grepl("UA",plot.df$sample),]

p2 <- ggplot(plot.df,aes(x=nodes,y=comp_size,color=sample))+
  geom_line(size=1)+
  labs(title="Degree-based attack",
       x="percentage of nodes removed",
       y="size of largest component",
       color="Network")+
  theme_bw()
```

### Randomness-based attacks
```{r}
plot.df <- as.data.frame(matrix(nrow=0,
                                ncol=3))

for(i in 1:length(net_spring)){
  g <- igraph::graph_from_adjacency_matrix(net_spring[[i]]$adjaMat1, weighted = TRUE)
  components <- igraph::clusters(g, mode="weak")
  print(max(components$csize))
  
  nodes <- length(V(g)$name)
  comp_size <- max(components$csize)
    
  while(length(V(g)$name) > 5){
    set.seed(123)
    g <- delete.vertices(g, V(g)$name[sample(1:length(V(g)$name), 1)])
    components <- igraph::clusters(g, mode="weak")
    
    nodes <- c(nodes,length(V(g)$name))
    comp_size <- c(comp_size,max(components$csize))
  }
  
  sample <- rep(paste0(names(net_spring)[i]," | ",covar.groups[1]),length(nodes))
  
  nodes <- (nodes[1]-nodes)/nodes[1]
  comp_size <- comp_size/comp_size[1]
  
  plot.df <- as.data.frame(rbind(plot.df,
                                 cbind(sample,
                                       nodes,
                                       comp_size)))
  #### 
  g <- igraph::graph_from_adjacency_matrix(net_spring[[i]]$adjaMat2, weighted = TRUE)
  components <- igraph::clusters(g, mode="weak")
  print(max(components$csize))
  
  nodes <- length(V(g)$name)
  comp_size <- max(components$csize)
    
  while(length(V(g)$name) > 5){
    set.seed(123)
    g <- delete.vertices(g, V(g)$name[sample(1:length(V(g)$name), 1)])
    components <- igraph::clusters(g, mode="weak")
    
    nodes <- c(nodes,length(V(g)$name))
    comp_size <- c(comp_size,max(components$csize))
  }
  
  sample <- rep(paste0(names(net_spring)[i]," | ",covar.groups[2]),length(nodes))
    
  nodes <- (nodes[1]-nodes)/nodes[1]
  comp_size <- comp_size/comp_size[1]
  
  plot.df <- as.data.frame(rbind(plot.df,
                                 cbind(sample,
                                       nodes,
                                       comp_size)))
}

plot.df$nodes <- as.numeric(as.character(plot.df$nodes))
plot.df$comp_size <- as.numeric(as.character(plot.df$comp_size))

plot.df <- plot.df[!grepl("UA",plot.df$sample),]
# plot.df <- plot.df[grepl("UA",plot.df$sample),]

p3 <- ggplot(plot.df,aes(x=nodes,y=comp_size,color=sample))+
  geom_line(size=1)+
  labs(title="Randomness-based attack",
       x="percentage of nodes removed",
       y="size of largest component",
       color="Network")+
  theme_bw()
```

### Plot attack-based plots

```{r}
p4 <- p1
p5 <- p2
p6 <- p3

```

```{r,fig.height=6,fig.width=12}
egg::ggarrange(p1+
                 guides(color='none'),
               p2+
                 guides(color='none'),
               p3,
               p4+
                 guides(color='none'),
               p5+
                 guides(color='none'),
               p6,
               labels=c("A","","","B","",""),
               nrow=2)sss

pdf(paste0(WORKING.DIR,"/plots/Fig6.pdf"),
    height=6,
    width=12,
    onefile = F)
egg::ggarrange(p1+
                 guides(color='none'),
               p2+
                 guides(color='none'),
               p3,
               p4+
                 guides(color='none'),
               p5+
                 guides(color='none'),
               p6,
               labels=c("A","","","B","",""),
               nrow=2)
dev.off()
```


