---
title: "Untitled"
output: html_document
---

# Preanalysis

## Load packages
```{r}
library(ConQuR)
library(doParallel)
library(dplyr)

WORKING.DIR="Z:/mchung/leopoldo_segal_07/"
```

# Set up metadata

This includes combining imsupp and nonimsupp samples, fixing sample names, and removing blood samples.

## Read in metadata
```{r}
runs <- list.files(paste0(WORKING.DIR,"/pipeline_output/"))

metadata.imsupp <- read.delim(paste0(WORKING.DIR,"/tables/w1to3_smpl_metadatamapping_covidpos_intub_imsupp_sequenced.csv"),
                              sep=",")
metadata.notimsupp <- read.delim(paste0(WORKING.DIR,"/tables/w1to3_smpl_metadatamapping_covidpos_intub_notimsupp_sequenced.csv"),
                                 row.names = 1,
                              sep=",")
```

## Fix metadata information

```{r}
metadata <- as.data.frame(rbind(metadata.imsupp[,intersect(colnames(metadata.imsupp),colnames(metadata.notimsupp))],
                                metadata.notimsupp[,intersect(colnames(metadata.imsupp),colnames(metadata.notimsupp))]))
metadata <- metadata[order(metadata$collection_date),]
metadata$Sample.ID[metadata$yl_sample_id == "UCS.0136.BKG.MainEvent"] <- "UCS.0136.BKG"
metadata <- metadata[metadata$sample_type != "Blood",] 
metadata$imsupp <-ifelse(metadata$Sample.ID %in% metadata.imsupp$Sample.ID,"Immunosuppressed","Healthy")

metadata$intub_to_sample_days <- as.numeric(as.character(metadata$intub_to_sample_days))
metadata <- metadata %>% mutate(weekpostintub=case_when(
                  intub_to_sample_days < 7 ~ 1,
                  intub_to_sample_days > 6 &  intub_to_sample_days < 14~ 2,
                  intub_to_sample_days > 13 &  intub_to_sample_days < 21~ 3,
                  intub_to_sample_days > 20 ~ 4))

metadata$run <- ifelse(metadata$run1_rna == 1,"run1",
                       ifelse(metadata$run2_rna == 1, "run2",
                              ifelse(metadata$run3_rna == 1,"run3","pilot")))
```

# Set up annotations

## Create annotation object

```{r}
annotation <- list()
```

## Read Bracken taxa information

```{r}
annotation$Bracken <- as.data.frame(matrix(nrow=0,ncol=0))
for(i in 1:length(runs)){
  file <- read.delim(paste0(WORKING.DIR,"/pipeline_output/",runs[i],"/tables/annotation.taxa.bracken.tsv"))
  annotation$Bracken <- as.data.frame(rbind(annotation$Bracken,file))
}
annotation$Bracken <- annotation$Bracken[!duplicated(annotation$Bracken$taxa),]
```

# Set up counts

## Create counts object

```{r}
counts <- list()
```

## Read in counts

### Read AMR gene counts for CARD

```{r}
rownames <- c()
for(i in 1:length(runs)){
  file <- read.delim(paste0(WORKING.DIR,"/pipeline_output/",runs[i],"/tables/counts.coverM.CARD.tsv"))
  rownames <- unique(c(rownames,rownames(file)))
}
counts$CARD <- as.data.frame(matrix(nrow=length(rownames),
                                    ncol=0))
rownames(counts$CARD) <- rownames
for(i in 1:length(runs)){
  file <- read.delim(paste0(WORKING.DIR,"/pipeline_output/",runs[i],"/tables/counts.coverM.CARD.tsv"))
  counts$CARD <- as.data.frame(cbind(counts$CARD,
                                     file[match(rownames(counts$CARD),rownames(file)),]))
}
```

### Read AMR gene counts for MEGARes

```{r}
rownames <- c()
for(i in 1:length(runs)){
  file <- read.delim(paste0(WORKING.DIR,"/pipeline_output/",runs[i],"/tables/counts.coverM.MEGARes.tsv"))
  rownames <- unique(c(rownames,rownames(file)))
}
counts$MEGARes <- as.data.frame(matrix(nrow=length(rownames),
                                    ncol=0))
rownames(counts$MEGARes) <- rownames
for(i in 1:length(runs)){
  file <- read.delim(paste0(WORKING.DIR,"/pipeline_output/",runs[i],"/tables/counts.coverM.MEGARes.tsv"))
  counts$MEGARes <- as.data.frame(cbind(counts$MEGARes,
                                     file[match(rownames(counts$MEGARes),rownames(file)),]))
}
```

### Read taxa gene counts for Bracken

```{r}
rownames <- c()
for(i in 1:length(runs)){
  file <- read.delim(paste0(WORKING.DIR,"/pipeline_output/",runs[i],"/tables/counts.taxa.Bracken.tsv"))
  rownames <- unique(c(rownames,rownames(file)))
}
counts$Bracken <- as.data.frame(matrix(nrow=length(rownames),
                                       ncol=0))
rownames(counts$Bracken) <- rownames
for(i in 1:length(runs)){
  file <- read.delim(paste0(WORKING.DIR,"/pipeline_output/",runs[i],"/tables/counts.taxa.Bracken.tsv"))
  counts$Bracken <- as.data.frame(cbind(counts$Bracken,
                                     file[match(rownames(counts$Bracken),rownames(file)),]))
}
counts$Bracken[is.na(counts$Bracken)] <- 0

counts$Bracken <- counts$Bracken[match(annotation$Bracken$taxa,rownames(counts$Bracken)),]
```

#### Edit SARS-CoV2 name

```{r}
rownames(counts$Bracken) <- gsub("Severe acute respiratory syndrome-related coronavirus","SARS-CoV2",rownames(counts$Bracken))
annotation$Bracken$taxa <- gsub("Severe acute respiratory syndrome-related coronavirus","SARS-CoV2",annotation$Bracken$taxa)
```

#### Remove human and protozoan entries

```{r}
bad_taxa <- c("Homo sapiens","Equine infectious anemia virus",
              annotation$Bracken$taxa[annotation$Bracken$superkingdom == "Eukaryota" &
                                      !is.na(annotation$Bracken$superkingdom) & 
                                      is.na(annotation$Bracken$kingdom)])

counts$Bracken <- counts$Bracken[!(rownames(counts$Bracken) %in% bad_taxa),]
annotation$Bracken <- annotation$Bracken[!(annotation$Bracken$taxa %in% bad_taxa),]
```

## Match counts samples to metadata samples
```{r}
counts <- lapply(counts,
                 function(x){
                  colnames(x) <- gsub("_S.*","",colnames(x))
                  x <- x[,metadata$Sample.ID]
                  return(x)
                 })
```

## Remove low count samples (<10000 counts)

```{r}
bad_samples <- colnames(counts$Bracken)[colSums(counts$Bracken) < 10000]
print(bad_samples)

metadata <- metadata[!(metadata$Sample.ID %in% bad_samples),]

counts <- lapply(counts,
                 function(x){
                  x <- x[,metadata$Sample.ID]
                  return(x)
                 })

```

# Correct counts batch effects using CONQUR

## Set up batch id covariates

```{r}
batchid <- factor(metadata$run,
                  levels=c("run1","run2","run3","pilot"))

# conqur.covar <- metadata[,c('subject_id','sample_type','adm_to_sample_days')]
conqur.covar <- metadata[,c('sample_type','adm_to_sample_days')]

conqur.covar$sample_type <- ifelse(conqur.covar$sample_type=='TA',
                                  1,
                                  ifelse(conqur.covar$sample_type=='BAL',
                                         2,
                                         ifelse(conqur.covar$sample_type=='UA',3,4)))
# conqur.covar$sex <- ifelse(conqur.covar$sex=='M',1,2)

```

## Run ConQuR
```{r}
counts.raw <- counts
for(i in 1:length(counts.raw)){
  counts[[i]] <- t(ConQuR(tax_tab=t(counts.raw[[i]]),
                          batchid=batchid, 
                          covariates=conqur.covar, 
                          batch_ref="run1"))
  print(i)
}
```

# Create R object
```{r}
save(metadata,
     annotation,
     counts,
     file=paste0(WORKING.DIR,"/upstream.R"))
```

